---
title: "ARTIS Data Ingest"
subtitle: "Changelog of raw data verions"
author: "Althea Marks"
format:
  pdf:
    papersize: letter
execute:
  echo: FALSE
  include: FALSE
  message: FALSE
  warning: FALSE
---

```{r setup}

#### Library #####
library(data.table)
library(dplyr)
library(readr)
library(stringr)
library(purrr)
library(tidyr)
library(knitr)
library(kableExtra)
library(readxl)


#### Directory setup #####

# Newest data
datadir_new <- file.path("~", "Documents", "UW-SAFS", "ARTIS", "data",
                         "model_inputs_raw_v2_0")
# last data ingest
datadir_old <- file.path("~", "Documents", "UW-SAFS", "ARTIS", "data",
                         "model_inputs_raw")
modelfundir <- file.path("~", "Documents", "git-projects", "artis-model", "R")


# values for report
raw_data_type <- "FAO Global Production"
artis_prev <- "v1.0"
artis_next <- "v2.0"
data_v_old <- "2022.1.1"
data_v_new <- "2025.1.0"

## FIXIT / NOTE: manually unzip downloaded files for this script to run properly
```

```{r get-files}

# List all subdirectories in datadir_new
subdirs_new <- list.dirs(datadir_new, recursive = TRUE, full.names = TRUE)
subdirs_old <- list.dirs(datadir_old, recursive = TRUE, full.names = TRUE)

# FAO Global production regex pattern
pattern_fao <- regex("(global[\\s_-]*production)", ignore_case = TRUE)

# FAO Global production directory
fao_dir_new <- subdirs_new[str_detect(subdirs_new, pattern_fao)]
fao_dir_old <- subdirs_old[str_detect(subdirs_old, pattern_fao)]

sp_new_fp <- file.path(
  fao_dir_new, 
  list.files(proddir_new, recursive = FALSE, full.names = FALSE) %>%
    grep("SPECIES_GROUPS.+", ., ignore.case = TRUE, value = TRUE))

sp_old_fp <- file.path(
  fao_dir_old, 
  list.files(proddir_old, recursive = FALSE, full.names = FALSE) %>%
    grep("SPECIES_GROUPS.+", ., ignore.case = TRUE, value = TRUE))

gp_new_fp <- file.path(fao_dir_new, "Global_production_quantity.csv")
gp_old_fp <- file.path(fao_dir_old, "Global_production_quantity.csv")

# BACI regex pattern
# pattern_baci <- regex("(baci)", ignore_case = TRUE)
# 
# # BACI directory
# baci_dir_new <- subdirs_new[str_detect(subdirs_new, pattern_baci)]
#baci_dir_old <- subdirs_old[str_detect(subdirs_old, pattern_baci)]

```

```{r assess-file-changes}

# 1. **Scan directories**
scan_dir <- function(dir, rg = "") {
  if (length(dir) == 0) return(character(0)) # Handle empty directories
  list.files(dir, pattern = rg, recursive = TRUE, full.names = TRUE)
}

# 2. **Read files from directories**
files_new <- scan_dir(fao_dir_new)
files_old <- scan_dir(fao_dir_old)

# 3. **Standardize file names**
all_files <- tibble(
  full_path = c(files_new, files_old),
  files = c(basename(files_new), basename(files_old)),
  directory = c(rep("new", length(files_new)), rep("old", length(files_old)))
) %>%
  mutate(
    files_std = files %>% 
      str_remove("^filtered_") %>%  # Remove "filtered_" prefix
      str_remove("_V\\d{6,7}[a-z]*") %>% # Remove version suffix (like _V202201)
      str_replace_all("\\.[a-zA-Z0-9]+$", "") %>%  # Remove extensions like .csv
      str_replace_all("[^a-zA-Z0-9]", "_") %>%   # Replace non-alphanumeric characters with underscores
      str_to_lower()  # Convert to lowercase
  )

# 4. **Pivot to match new and old files**
file_level_log <- all_files %>%
  pivot_wider(
    id_cols = files_std,             # Only use files_std to identify matching files
    names_from = directory, 
    values_from = full_path, 
    names_prefix = ""
  ) %>%
  mutate(
    exists_in_old = !is.na(old),
    exists_in_new = !is.na(new),
    size_old_mb = ifelse(!is.na(old), 
                         file.info(old)$size / 1e6, 
                         NA_real_),
    size_new_mb = ifelse(!is.na(new), 
                         file.info(new)$size / 1e6, 
                         NA_real_),
    size_change_mb = size_new_mb - size_old_mb
  )

# 5. **Data-Level Change Log**
comparison_pairs <- file_level_log %>%
  filter(exists_in_old & exists_in_new) %>%
  # compare matching .csv files 
  filter((str_detect(new, "\\.csv$") & str_detect(old, "\\.csv$")) |
           # OR matching .xlsx files
           (str_detect(new, "\\.xlsx$") & str_detect(old, "\\.xlsx$"))) %>%
  select(files_std, new, old)

library(readxl)

compare_files <- function(file_new, file_old) {
  if(basename(file_new) %>% str_detect("\\.csv$")) {
    data_new <- fread(file_new)
  } 
  if(basename(file_new) %>% str_detect("\\.xlsx$")) {
    data_new <- readxl::read_xlsx(file_new, skip=1)
  }
  if(basename(file_old) %>% str_detect("\\.csv$")) {
    data_old <- fread(file_old)
  }
  if(basename(file_old) %>% str_detect("\\.xlsx$")) {
    data_old <- readxl::read_xlsx(file_old, skip=1)
  }

  
  list(
    file_name_new = basename(file_new),
    file_name_old = basename(file_old),
    new_rows = nrow(data_new),
    old_rows = nrow(data_old),
    row_diff = nrow(data_new) - nrow(data_old),
    new_cols = colnames(data_new),
    old_cols = colnames(data_old),
    # single col value will create a vector column - >1 cols will create list-col
    added_cols = if (length(setdiff(colnames(data_new),
                                    colnames(data_old))) > 0) {
      setdiff(colnames(data_new), colnames(data_old)) 
    } else { NA_character_},
    removed_cols = if (length(setdiff(colnames(data_old),
                                      colnames(data_new))) > 0) {
      setdiff(colnames(data_old), colnames(data_new)) 
    } else {NA_character_},
                   
    data_type_changes = tibble(
      column = union(colnames(data_new), colnames(data_old)),
      # if col is in data_old get its class - else it is NA_character_
      old_type = map_chr(union(colnames(data_new), colnames(data_old)), 
                         ~ ifelse(.x %in% colnames(data_old), 
                                  class(data_old[[.x]])[1], 
                                  NA_character_)),
      # if col is in data_new get its class - else it is NA_character_
      new_type = map_chr(union(colnames(data_new), 
                               colnames(data_old)), 
                         ~ ifelse(.x %in% colnames(data_new), 
                                  class(data_new[[.x]])[1], 
                                  NA_character_))
    ) %>% 
      mutate(type_changed = old_type != new_type)
  )
}

data_level_log <- comparison_pairs %>%
  mutate(comparison = map2(new, old, ~ compare_files(.x, .y)))

# Extract data-level comparison details
data_change_log <- data_level_log %>%
  unnest_wider(comparison) %>%
  select(file_name_new, file_name_old, new_rows, old_rows, row_diff, 
         added_cols, removed_cols, data_type_changes)


```

# Summary

The purpose of this report is to examine and document changes 
between raw input data versions ingested into the ARTIS (Aquatic Resource Trade 
in Species) Model. These raw data will then be processed and standardized to the 
model inputs (data build) fed directly into the ARTIS model. 

### ARTIS model versions

- ARTIS ``r artis_prev`` was the previous model run
- ARTIS ``r artis_next`` is upcoming model run

# FAO Global Production

### Raw input data versions

- `r raw_data_type` ``r data_v_old`` is the previous data version
- `r raw_data_type` ``r data_v_new`` is the data version to be ingested in the upcoming ARTIS ``r artis_next`` model run

### File changes

```{r logic-files-missing}
#| results: 'asis'
#| include: TRUE

if (any(file_level_log$exists_in_old == FALSE)) {
  cat(paste0("The new ", data_v_new, " data version contains files that are not present in the old ", data_v_old, " version: ", basename(file_level_log[[which(file_level_log$exists_in_old == FALSE), "new"]])
                 )
          )
}
```

```{r logic-files-added}
#| results: 'asis'
#| include: TRUE

if (any(file_level_log$exists_in_new == FALSE)) {
  cat(paste0("The old ", data_v_old, " data version contains files that are not present in the new ", data_v_new, " version: ", basename(file_level_log[[which(file_level_log$exists_in_new == FALSE), "old"]])
                 )
          )
}
```

### Column changes

Only matching .csv and .xlsx files were compared.

```{r}
#| label: tab:added_removed_columns
#| include: TRUE
#| results: 'asis'

# Filter the data_change_log to only include files with added or removed columns
added_cols_log <- data_change_log %>% 
  filter(!is.na(added_cols) | !is.na(removed_cols)) %>% 
  select(file_name_new, added_cols, removed_cols)

if (nrow(added_cols_log) > 0 ) {
  # Create a kableExtra table with automatic scaling
  added_cols_log %>%
    kbl(
      col.names = c("File Name", "Added Columns", "Removed Columns"),
      caption = "Files where columns were added/removed in the new version.",
      align = c('l', 'c', 'c')
    ) %>%
    kable_styling(
      full_width = TRUE, 
      latex_options = c("scale_down", "hold_position")
    )
}

```

### Country and species

```{r}

##### Changes in species using prod

# Need to use production to assess species, SPECIES_GROUPS has many more species not reported in prod
sp_old <- fread(sp_old_fp)
gp_old <- fread(gp_old_fp)

gp_sp_old <- gp_old |> 
  select(SPECIES.ALPHA_3_CODE) |> 
  left_join(
    sp_old |> 
      rename(SPECIES.ALPHA_3_CODE = `3A_Code`) |> 
      select(SPECIES.ALPHA_3_CODE, Scientific_Name),
    by = "SPECIES.ALPHA_3_CODE") |> 
      distinct()

sp_new <- fread(sp_new_fp)
gp_new <- fread(gp_new_fp)

gp_sp_new <- gp_new |> 
  select(SPECIES.ALPHA_3_CODE) |> 
  left_join(
    sp_new |> 
      rename(SPECIES.ALPHA_3_CODE = `3A_Code`) |> 
      select(SPECIES.ALPHA_3_CODE, Scientific_Name),
    by = "SPECIES.ALPHA_3_CODE") |> 
      distinct()

sp_comp <- gp_sp_new |> 
  mutate(present_new = 1) |> 
  full_join(
    gp_sp_old |> 
      mutate(present_old = 1),
      by = c("SPECIES.ALPHA_3_CODE", "Scientific_Name")) |>
  filter(is.na(present_new) | is.na(present_old)) 

sp_comp  |> group_by(SPECIES.ALPHA_3_CODE)  |> tally()  |> filter(n == 1)  |> nrow()

###### Change in countries 
country_n_old <- length(unique(prod_old$country_iso3_alpha))
country_n_new <- length(unique(prod_new$country_iso3_alpha))

country_unique_old <- setdiff(unique(prod_old$country_iso3_alpha),
                              unique(prod_new$country_iso3_alpha))
country_unique_new <- setdiff(unique(prod_new$country_iso3_alpha),
                              unique(prod_old$country_iso3_alpha))

# countries_identical <- any(sort(unique(prod_old$country_iso3_alpha)) == 
#                              sort(unique(prod_new$country_iso3_alpha)))
```

```{r country-species-changes}
# source function to build usable 2022 FAO production data
source(file.path(modelfundir, "rebuild_fao_2022_dat.R"))
# source function to build usable 2024 FAO production data
source(file.path(modelfundir, "rebuild_fao_2024_dat.R"))

# combine and organize prod files
prod_old <- rebuild_fao_2022_dat(datadir = datadir_old, 
                                 filename = "GlobalProduction_2022.1.1.zip")
prod_new <- rebuild_fao_2024_dat(datadir = datadir_new, 
                                 filename = "GlobalProduction_2024.1.0.zip")

# define cleaning function
# code is copied and changed from `classify_prod_dat.R` function in artis-model repo
clean_prod <- function(a_prod_file, old_or_new = "old") {
  
  cols_keep <- c("country", "prod_method", "species_name_en", 
                 "species_scientific_name", "country_iso3_code", 
                 "species_major_group", "yearbook_group_en")
  # code is copied and changed from `classify_prod_dat.R` function in artis-model repo
      prod_ts <- a_prod_file %>%
        # select columns we are interested in for the purposes of comparing version changes
        select(any_of(cols_keep)) %>% 
      # Standardize column names between FAO and SAU datasets 
      dplyr::rename(
        CommonName = species_name_en, 
        SciName = species_scientific_name,
        country_iso3_alpha = country_iso3_code, # alpha iso code
        country_iso3_numeric = country) %>% # numeric iso code 
      mutate(CommonName=tolower(as.character(CommonName)),
             SciName=tolower(as.character(SciName))) %>%
      # Trim any leading/trailing whitespace
      mutate_all(str_trim) %>%
      { if(old_or_new == "old") {   
      # Filter out groups not considered in this analysis  
      filter(., !species_major_group %in% c("PLANTAE AQUATICAE",
                                         "MAMMALIA",
                                         "Crocodiles and alligators",
                                         "AMPHIBIA, REPTILIA",
                                         "Turtles",
                                         "Frogs and other amphibians",
                                         "Corals",
                                         "Sponges", 
                                         "Pearls, mother-of-pearl, shells")) 
        } else if (old_or_new == "new") {
              # Filter out groups not considered in this analysis  
      filter(., !species_major_group %in% c("PLANTAE AQUATICAE",
                                        "AMPHIBIA, REPTILIA",
                                        "MAMMALIA"),
             # includes corals, sponges, pearl oysters, shells 
             !yearbook_group_en == "Other aq. animals & products")
          }
      } %>%
      mutate(
        SciName = gsub(SciName, 
                       pattern=" \\(\\=.*", 
                       replacement=""))
      
      return(prod_ts)
  
}

# clean prod versions 
prod_old <- clean_prod(prod_old, old_or_new = "old")
prod_new <- clean_prod(prod_new, old_or_new = "new")

###### Change in countries 
country_n_old <- length(unique(prod_old$country_iso3_alpha))
country_n_new <- length(unique(prod_new$country_iso3_alpha))

country_unique_old <- setdiff(unique(prod_old$country_iso3_alpha),
                              unique(prod_new$country_iso3_alpha))
country_unique_new <- setdiff(unique(prod_new$country_iso3_alpha),
                              unique(prod_old$country_iso3_alpha))

# countries_identical <- any(sort(unique(prod_old$country_iso3_alpha)) == 
#                              sort(unique(prod_new$country_iso3_alpha)))

####### Change in species 
# before taxa cleaning, matching, and fishbase merging (classify_prod_dat.R)
sp_n_old <- length(unique(prod_old$SciName))
sp_n_new <- length(unique(prod_new$SciName))

sp_unique_old <- setdiff(unique(prod_old$SciName),
                         unique(prod_new$SciName))
sp_unique_new <- setdiff(unique(prod_new$SciName),
                         unique(prod_old$SciName))

```

- Countries in the FAO Global Production data that are different:
  - unique to the old version: `r country_unique_old`
  - unique to the new version: `r country_unique_new`

- Species in the FAO Global Production data that are different:  
  - unique to the old version: 

```{r}
#| include: true
#| output: true

sort(sp_unique_old)
```

  - unique to the new version: 

```{r}
#| include: true
#| output: true

sort(sp_unique_new)
```

