---
title: "ARTIS Raw Data Assessment"
subtitle: "Changelog of raw data versions"
author: "Althea Marks"
date: today
format:
  pdf:
    papersize: letter
    toc: true
    toc-depth: 2
    number-sections: true
execute:
  echo: FALSE
  include: FALSE
  message: FALSE
  warning: FALSE
---

```{r setup}
#### Library #####
{
  library(data.table)
  library(dplyr)
  library(readr)
  library(stringr)
  library(purrr)
  library(tidyr)
  library(knitr)
  library(kableExtra)
  library(readxl)
  library(gt)
  library(artis)
  library(glue)
  library(testthat)
}

#### Directory setup #####

# Newest data
datadir_new <- file.path("~", "Documents", "UW-SAFS", "ARTIS", "data",
                         "model_inputs_raw_3.0")
# Last data ingest
datadir_old <- file.path("~", "Documents", "UW-SAFS", "ARTIS", "data",
                         "model_inputs_raw_2.0")

# FishBase / SeaLifeBase directories
fb_slb_dir_new <- file.path(datadir_new, "fishbase_sealifebase_25.04")
fb_slb_dir_old <- file.path(datadir_old, "fishbase_sealifebase_24.07")

# Report version labels
raw_data_type <- "FAO Global Production"
artis_old     <- "2.0"
artis_new     <- "3.0"
data_v_old    <- "2025.1.0"
data_v_new    <- "2026.1.0"
baci_v_old    <- "V202501"
baci_v_new    <- "V202601"

# Status emoji constants
STATUS_PASS   <- "\u2705 PASS"
STATUS_NOTE   <- "\U0001F4CB NOTE"
STATUS_REVIEW <- "\U0001F50D REVIEW"
STATUS_FAIL   <- "\u274C FAIL"

# Per-section validation logs
fao_log   <- list()
fbslb_log <- list()
baci_log  <- list()
```

```{r helper-fns}
render_validation_table <- function(log, title = "Validation Results") {
  if (length(log) == 0) {
    return(
      gt::gt(data.frame(Check = character(), Status = character(), Detail = character())) |>
        gt::tab_header(title = title)
    )
  }

  tbl <- log |>
    tibble::enframe(name = "check", value = "result") |>
    mutate(
      status = map_chr(result, "status"),
      detail = map_chr(result, "detail")
    ) |>
    select(check, status, detail)

  tbl |>
    gt::gt() |>
    gt::tab_header(title = title) |>
    gt::cols_label(
      check  = "Check",
      status = "Status",
      detail = "Detail"
    ) |>
    gt::tab_style(
      style     = gt::cell_fill(color = "#d9ead3"),
      locations = gt::cells_body(rows = str_detect(status, "PASS"))
    ) |>
    gt::tab_style(
      style     = gt::cell_fill(color = "#fff2cc"),
      locations = gt::cells_body(rows = str_detect(status, "NOTE"))
    ) |>
    gt::tab_style(
      style     = gt::cell_fill(color = "#fce5cd"),
      locations = gt::cells_body(rows = str_detect(status, "REVIEW"))
    ) |>
    gt::tab_style(
      style     = gt::cell_fill(color = "#f4cccc"),
      locations = gt::cells_body(rows = str_detect(status, "FAIL"))
    ) |>
    gt::cols_width(
      check  ~ gt::px(220),
      status ~ gt::px(100),
      detail ~ gt::px(300)
    ) |>
    gt::tab_options(table.font.size = 10)
}
```

# Summary

The purpose of this report is to examine and document changes between raw input
data versions ingested into the ARTIS (Aquatic Resource Trade in Species) Model.
These raw data will then be processed and standardized to the model inputs (data
build) fed directly into the ARTIS model.

::: {.callout-note}
## Status Code Legend

Each validation check is assigned one of four status codes:

- **`r STATUS_PASS`** --- No change detected; no action needed.
- **`r STATUS_NOTE`** --- Expected or informational change (e.g., row count increase, year coverage extension, new taxa added). Review is encouraged but no code changes are required.
- **`r STATUS_REVIEW`** --- Change needs attention before the model run (e.g., new taxa in FAO data may require updates to `classify_prod_dat()` in `01_clean_input_data.R`, dropped countries, trade value/quantity changes, product codes removed).
- **`r STATUS_FAIL`** --- Blocking issue (e.g., missing files, missing key columns, year regression, coverage drop). Must be resolved before proceeding.
:::

### ARTIS model versions

- ARTIS `r artis_old` was the previous model run
- ARTIS `r artis_new` is the upcoming model run

# FAO Global Production

Raw input data versions compared:

- `r raw_data_type` `r data_v_old` --- previous version (ARTIS `r artis_old`)
- `r raw_data_type` `r data_v_new` --- version to be ingested for ARTIS `r artis_new`

The checks below assess file presence, column structure, row counts, year
coverage, total production quantities, and species coverage between the two
data versions. Results are summarised in the validation table at the end of
this section.

```{r fao-get-dirs}
# List all subdirectories
subdirs_new <- list.dirs(datadir_new, recursive = TRUE, full.names = TRUE)
subdirs_old <- list.dirs(datadir_old, recursive = TRUE, full.names = TRUE)

# FAO Global production regex pattern
pattern_fao <- regex("(global[\\s_-]*production)", ignore_case = TRUE)

# FAO Global production directories
fao_dir_new <- subdirs_new[str_detect(subdirs_new, pattern_fao)]
fao_dir_old <- subdirs_old[str_detect(subdirs_old, pattern_fao)]
```

```{r fao-scan-files}
scan_dir <- function(dir, rg = "") {
  if (length(dir) == 0) return(character(0))
  list.files(dir, pattern = rg, recursive = TRUE, full.names = TRUE)
}

files_new <- scan_dir(fao_dir_new)
files_old <- scan_dir(fao_dir_old)

# Standardise file names for matching
all_files <- tibble(
  full_path = c(files_new, files_old),
  files     = c(basename(files_new), basename(files_old)),
  directory = c(rep("new", length(files_new)), rep("old", length(files_old)))
) |>
  mutate(
    files_std = files |>
      str_remove("^filtered_") |>
      str_remove("_V\\d{6,7}[a-z]*") |>
      str_replace_all("\\.[a-zA-Z0-9]+$", "") |>
      str_replace_all("[^a-zA-Z0-9]", "_") |>
      str_to_lower()
  )

file_level_log <- all_files |>
  pivot_wider(
    id_cols      = files_std,
    names_from   = directory,
    values_from  = full_path,
    names_prefix = ""
  ) |>
  mutate(
    exists_in_old  = !is.na(old),
    exists_in_new  = !is.na(new),
    size_old_mb    = ifelse(!is.na(old), file.info(old)$size / 1e6, NA_real_),
    size_new_mb    = ifelse(!is.na(new), file.info(new)$size / 1e6, NA_real_),
    size_change_mb = size_new_mb - size_old_mb
  )
```

```{r fao-log-file-presence}
files_only_in_new <- file_level_log |> filter(!exists_in_old)
files_only_in_old <- file_level_log |> filter(!exists_in_new)

if (nrow(files_only_in_new) == 0 && nrow(files_only_in_old) == 0) {
  fao_log[["FAO file presence"]] <- list(
    status = STATUS_PASS,
    detail = "All files present in both versions."
  )
} else {
  if (nrow(files_only_in_new) > 0) {
    added_names <- basename(files_only_in_new$new) |> paste(collapse = ", ")
    fao_log[[glue("FAO files added in {data_v_new}")]] <- list(
      status = STATUS_NOTE,
      detail = glue("Files present in {data_v_new} but not {data_v_old}: {added_names}")
    )
  }
  if (nrow(files_only_in_old) > 0) {
    removed_names <- basename(files_only_in_old$old) |> paste(collapse = ", ")
    fao_log[[glue("FAO files removed (not in {data_v_new})")]] <- list(
      status = STATUS_NOTE,
      detail = glue("Files present in {data_v_old} but not {data_v_new}: {removed_names}")
    )
  }
}
```

```{r fao-column-row-changes}
compare_files <- function(file_new, file_old) {
  if (basename(file_new) |> str_detect("\\.csv$")) {
    data_new <- fread(file_new)
  }
  if (basename(file_new) |> str_detect("\\.xlsx$")) {
    data_new <- readxl::read_xlsx(file_new, skip = 1)
  }
  if (basename(file_old) |> str_detect("\\.csv$")) {
    data_old <- fread(file_old)
  }
  if (basename(file_old) |> str_detect("\\.xlsx$")) {
    data_old <- readxl::read_xlsx(file_old, skip = 1)
  }

  list(
    file_name_new = basename(file_new),
    file_name_old = basename(file_old),
    new_rows      = nrow(data_new),
    old_rows      = nrow(data_old),
    row_diff      = nrow(data_new) - nrow(data_old),
    new_cols      = colnames(data_new),
    old_cols      = colnames(data_old),
    added_cols    = if (length(setdiff(colnames(data_new), colnames(data_old))) > 0) {
      setdiff(colnames(data_new), colnames(data_old))
    } else { NA_character_ },
    removed_cols  = if (length(setdiff(colnames(data_old), colnames(data_new))) > 0) {
      setdiff(colnames(data_old), colnames(data_new))
    } else { NA_character_ },
    data_type_changes = tibble(
      column   = union(colnames(data_new), colnames(data_old)),
      old_type = map_chr(union(colnames(data_new), colnames(data_old)),
                         ~ ifelse(.x %in% colnames(data_old), class(data_old[[.x]])[1], NA_character_)),
      new_type = map_chr(union(colnames(data_new), colnames(data_old)),
                         ~ ifelse(.x %in% colnames(data_new), class(data_new[[.x]])[1], NA_character_))
    ) |>
      mutate(type_changed = old_type != new_type)
  )
}

comparison_pairs <- file_level_log |>
  filter(exists_in_old & exists_in_new) |>
  filter(
    (str_detect(new, "\\.csv$") & str_detect(old, "\\.csv$")) |
    (str_detect(new, "\\.xlsx$") & str_detect(old, "\\.xlsx$"))
  ) |>
  select(files_std, new, old)

data_level_log <- comparison_pairs |>
  mutate(comparison = map2(new, old, ~ compare_files(.x, .y)))

data_change_log <- data_level_log |>
  unnest_wider(comparison) |>
  select(file_name_new, file_name_old, new_rows, old_rows, row_diff,
         added_cols, removed_cols, data_type_changes)

# Column change logs
col_changes <- data_change_log |>
  filter(!is.na(added_cols) | !is.na(removed_cols))

if (nrow(col_changes) == 0) {
  fao_log[["FAO column changes"]] <- list(
    status = STATUS_PASS,
    detail = "No column additions or removals detected across matched files."
  )
} else {
  walk(seq_len(nrow(col_changes)), function(i) {
    row     <- col_changes[i, ]
    fname   <- row$file_name_new
    added   <- if (!is.na(row$added_cols))   paste(unlist(row$added_cols),   collapse = ", ") else "none"
    removed <- if (!is.na(row$removed_cols)) paste(unlist(row$removed_cols), collapse = ", ") else "none"
    fao_log[[glue("FAO column changes: {fname}")]] <<- list(
      status = STATUS_NOTE,
      detail = glue("Added: {added}. Removed: {removed}.")
    )
  })
}

# Row count logs
walk(seq_len(nrow(data_change_log)), function(i) {
  row   <- data_change_log[i, ]
  fname <- row$file_name_new
  fao_log[[glue("FAO row count: {fname}")]] <<- list(
    status = STATUS_NOTE,
    detail = glue(
      "{data_v_old}: {row$old_rows} rows | {data_v_new}: {row$new_rows} rows | diff: {row$row_diff}"
    )
  )
})
```

```{r fao-year-coverage}
raw_prod_old_fn <- glue("GlobalProduction_{data_v_old}.zip")
raw_prod_new_fn <- glue("GlobalProduction_{data_v_new}.zip")

fao_prod_old <- artis::rebuild_fao_2023_dat(
  datadir  = datadir_old,
  filename = raw_prod_old_fn
)

fao_prod_new <- artis::rebuild_fao_2023_dat(
  datadir  = datadir_new,
  filename = raw_prod_new_fn
)

min_yr_old <- min(fao_prod_old$year, na.rm = TRUE)
max_yr_old <- max(fao_prod_old$year, na.rm = TRUE)
min_yr_new <- min(fao_prod_new$year, na.rm = TRUE)
max_yr_new <- max(fao_prod_new$year, na.rm = TRUE)

# Check min year <= 1996 (old)
fao_log[["FAO year coverage: min year (old)"]] <- list(
  status = if (min_yr_old > 1996) STATUS_FAIL else STATUS_PASS,
  detail = if (min_yr_old > 1996) {
    glue("Old data min year is {min_yr_old}; expected <= 1996.")
  } else {
    glue("Old data min year: {min_yr_old} (<= 1996). OK.")
  }
)

# Check min year <= 1996 (new)
fao_log[["FAO year coverage: min year (new)"]] <- list(
  status = if (min_yr_new > 1996) STATUS_FAIL else STATUS_PASS,
  detail = if (min_yr_new > 1996) {
    glue("New data min year is {min_yr_new}; expected <= 1996.")
  } else {
    glue("New data min year: {min_yr_new} (<= 1996). OK.")
  }
)

# Check max year progression
fao_log[["FAO year coverage: max year"]] <- list(
  status = if (max_yr_new < max_yr_old) STATUS_FAIL else STATUS_NOTE,
  detail = if (max_yr_new < max_yr_old) {
    glue("New data max year ({max_yr_new}) is less than old ({max_yr_old}). Regression detected.")
  } else {
    glue("Max year: {data_v_old} = {max_yr_old}, {data_v_new} = {max_yr_new}. Expected extension.")
  }
)
```

```{r fao-production-quantity}
qty_old <- fao_prod_old |>
  filter(year >= 1996) |>
  group_by(year) |>
  summarise(total_qty_old = sum(quantity, na.rm = TRUE), .groups = "drop")

qty_new <- fao_prod_new |>
  filter(year >= 1996) |>
  group_by(year) |>
  summarise(total_qty_new = sum(quantity, na.rm = TRUE), .groups = "drop")

qty_compare <- full_join(qty_old, qty_new, by = "year") |>
  mutate(
    qty_diff     = total_qty_new - total_qty_old,
    qty_diff_pct = round(100 * qty_diff / total_qty_old, 2)
  ) |>
  arrange(year)

qty_changed <- qty_compare |> filter(!is.na(qty_diff) & qty_diff != 0)

fao_log[["FAO total production quantity by year"]] <- list(
  status = if (nrow(qty_changed) == 0) STATUS_PASS else STATUS_NOTE,
  detail = if (nrow(qty_changed) == 0) {
    "Total production quantity identical across all shared years (1996+)."
  } else {
    glue("{nrow(qty_changed)} year(s) show quantity differences. See detail table below.")
  }
)
```

```{r fao-sciname-diff}
raw_prod_old_sp <- fao_prod_old |>
  select(sciname = species_scientific_name, common_name = species_name_en) |>
  mutate(sciname = tolower(sciname), common_name = tolower(common_name)) |>
  distinct()

raw_prod_new_sp <- fao_prod_new |>
  select(sciname = species_scientific_name, common_name = species_name_en) |>
  mutate(sciname = tolower(sciname), common_name = tolower(common_name)) |>
  distinct()

raw_prod_old_sp_only <- anti_join(raw_prod_old_sp, raw_prod_new_sp, by = "sciname") |>
  arrange(sciname) |>
  mutate(version = data_v_old)

raw_prod_new_sp_only <- anti_join(raw_prod_new_sp, raw_prod_old_sp, by = "sciname") |>
  arrange(sciname) |>
  mutate(version = data_v_new)

if (nrow(raw_prod_new_sp_only) == 0 && nrow(raw_prod_old_sp_only) == 0) {
  fao_log[["FAO sciname: taxa unique to new version"]] <- list(
    status = STATUS_PASS,
    detail = "No taxa differences between versions."
  )
} else {
  fao_log[["FAO sciname: taxa unique to new version"]] <- list(
    status = if (nrow(raw_prod_new_sp_only) > 0) STATUS_REVIEW else STATUS_PASS,
    detail = if (nrow(raw_prod_new_sp_only) > 0) {
      glue(
        "{nrow(raw_prod_new_sp_only)} taxa appear only in {data_v_new}. ",
        "May require updates to classify_prod_dat() in 01_clean_input_data.R. ",
        "Some will be resolved via synonym matching; others may need manual case_when corrections."
      )
    } else {
      glue("No taxa unique to {data_v_new}.")
    }
  )

  fao_log[["FAO sciname: taxa dropped from old version"]] <- list(
    status = if (nrow(raw_prod_old_sp_only) > 0) STATUS_NOTE else STATUS_PASS,
    detail = if (nrow(raw_prod_old_sp_only) > 0) {
      glue(
        "{nrow(raw_prod_old_sp_only)} taxa appear only in {data_v_old}. ",
        "Expected attrition; unlikely to require code changes."
      )
    } else {
      glue("No taxa unique to {data_v_old}.")
    }
  )
}
```

### Column changes

Only matching .csv and .xlsx files were compared.

```{r tab-added-removed-columns}
#| include: TRUE
#| results: 'asis'

added_cols_log <- data_change_log |>
  filter(!is.na(added_cols) | !is.na(removed_cols)) |>
  select(file_name_new, added_cols, removed_cols)

if (nrow(added_cols_log) > 0) {
  added_cols_log |>
    kbl(
      col.names = c("File Name", "Added Columns", "Removed Columns"),
      caption   = "Files where columns were added/removed in the new version.",
      align     = c("l", "c", "c")
    ) |>
    kable_styling(
      full_width    = TRUE,
      latex_options = c("scale_down", "hold_position")
    )
}
```

### Production quantity changes by year

```{r tab-qty-changes}
#| include: TRUE
#| results: 'asis'

if (nrow(qty_changed) > 0) {
  qty_changed |>
    kbl(
      col.names = c(
        "Year",
        glue("Total Qty ({data_v_old})"),
        glue("Total Qty ({data_v_new})"),
        "Diff", "% Diff"
      ),
      caption = "Years where total FAO production quantity differs between versions (1996+).",
      digits  = 0
    ) |>
    kable_styling(
      full_width    = TRUE,
      latex_options = c("scale_down", "hold_position")
    )
}
```

### Sciname differences

```{r tab-sciname-new}
#| include: TRUE
#| results: 'asis'

raw_prod_new_sp_only |>
  kable(
    caption   = glue("Taxa unique to {data_v_new} FAO production data"),
    booktabs  = TRUE,
    col.names = c("Scientific Name", "Common Name", "Version")
  ) |>
  kable_styling(font_size = 8)
```

```{r tab-sciname-old}
#| include: TRUE
#| results: 'asis'

raw_prod_old_sp_only |>
  kable(
    caption   = glue("Taxa unique to {data_v_old} FAO production data"),
    booktabs  = TRUE,
    col.names = c("Scientific Name", "Common Name", "Version")
  ) |>
  kable_styling(font_size = 8)
```

### FAO Validation Summary

```{r fao-validation-table}
#| include: TRUE

render_validation_table(
  fao_log,
  title = glue("FAO Global Production Validation: {data_v_old} vs {data_v_new}")
)
```
