---
title: "ARTIS Raw Data Assessment"
subtitle: "Assumption checks for ingesting raw external data"
author: "Althea Marks"
date: today
output: html_document
execute:
  echo: FALSE
  include: FALSE
  message: FALSE
  warning: FALSE
---

```{r setup}
{
  library(data.table)
  library(dplyr)
  library(readr)
  library(stringr)
  library(purrr)
  library(tidyr)
  library(knitr)
  library(kableExtra)
  library(readxl)
  library(gt)
  library(artis)
  library(glue)
  library(readxl)
}


#### Directory setup #####
# ARTIS versions ingesting new data
artis_old <- "2.0"
artis_new <- "3.0"

path_data <- file.path("~/Documents/UW-SAFS/ARTIS/data")

path_raw_data_old <- file.path(path_data, glue("model_inputs_raw_{artis_old}"))
path_raw_data_new <- file.path(path_data, glue("model_inputs_raw_{artis_new}"))

# FAO production
fao_v_old <- "2025.1.0"
fao_v_new <- "2026.1.0"
# BACI trade
baci_v_old <- "V202501"
baci_v_new <- "V202601"
# Fishbase / Sealifebase
fbslb_v_old <- "24.07"
fbslb_v_new <- "25.04"

# List all subdirectories in path_raw_data_*
subdirs_new <- list.dirs(path_raw_data_new, recursive = TRUE, full.names = TRUE)
subdirs_old <- list.dirs(path_raw_data_old, recursive = TRUE, full.names = TRUE)

# FAO Prod assessment
# Threshold: flag rows where production absolute % change exceeds this value
qty_pct_threshold <- 1  # percent
```

# Summary

The purpose of this report is to examine and document changes between raw input
data versions ingested into the ARTIS (Aquatic Resource Trade in Species) Model.
These raw data will then be cleaned and standardized to generate the model inputs fed directly into the ARTIS model.

Changes to the raw data often requires updates or tweaks to the `01_clean_input_data.R` script used to generate the model inputs. Use the summary tables to understand what may need adjustments in the code. 

## Model and Data Versions

### ARTIS model versions

- Previous ARTIS: ``r artis_old``
- Upcoming ARTIS: ``r artis_new``

### FAO Global Production

- Previous: ``r fao_v_old``
- Incoming: ``r fao_v_new``

### BACI International Bilateral Trade

- Previous: ``r baci_v_old``
- Incoming: ``r baci_v_new``

### Fishbase / Sealifebase

- Previous: ``r fbslb_v_old``
- Incoming: ``r fbslb_v_new``

# Assess: FAO Global Production

## Files present in dataset

### Check that ... 

- The data set includes files required to run `artis::rebuild_fao_yyyy.R` in `01-clean-input-data.R`
  - Green rows = files new in the incoming ``r fao_v_new`` version. 
  - Red rows = files removed from the ``r fao_v_new`` incoming version.
- The file sizes and row remain the same, **EXCEPT** `CL_FI_SPECIES_GROUPS` and ` Global_production_quantity` that are expected to generally increase in size and row with version bumps. 

```{r fao-get-files}

# FAO Global production regex pattern
fao_pattern <- regex("(global[\\s_-]*production)", ignore_case = TRUE)

# FAO Global production directory
fao_dir_new <- subdirs_new[str_detect(subdirs_new, fao_pattern)]
fao_dir_old <- subdirs_old[str_detect(subdirs_old, fao_pattern)]

fao_gp_new <- file.path(fao_dir_new, "Global_production_quantity.csv")
fao_gp_old <- file.path(fao_dir_old, "Global_production_quantity.csv")

# BACI regex pattern
# pattern_baci <- regex("(baci)", ignore_case = TRUE)
# 
# # BACI directory
# baci_dir_new <- subdirs_new[str_detect(subdirs_new, pattern_baci)]
#baci_dir_old <- subdirs_old[str_detect(subdirs_old, pattern_baci)]

```

```{r fao-scan-files}

# 1. **Scan directories**
# Returns full file paths from a directory; rg allows optional regex filtering
scan_dir <- function(dir, rg = "") {
  if (length(dir) == 0) return(character(0)) # Handle empty directories
  list.files(dir, pattern = rg, recursive = TRUE, full.names = TRUE)
}

# 2. **Read files from directories**
files_new <- scan_dir(fao_dir_new)
files_old <- scan_dir(fao_dir_old)

# 3. **Standardize file names**
# Combine new and old file paths into a single tibble with a directory label,
# then derive files_std: a version-agnostic, lowercase, underscore-separated key
# used to match the same logical file across old and new versions
all_files <- tibble(
  full_path = c(files_new, files_old),
  files = c(basename(files_new), basename(files_old)),
  directory = c(rep("new", length(files_new)), rep("old", length(files_old)))
) %>%
  mutate(
    files_std = files %>% 
      str_remove("^filtered_") %>%  # Remove "filtered_" prefix
      str_remove("_V\\d{6,7}[a-z]*") %>% # Remove version suffix (like _V202201)
      str_replace_all("\\.[a-zA-Z0-9]+$", "") %>%  # Remove extensions like .csv
      str_replace_all("[^a-zA-Z0-9]", "_") %>%   # Replace non-alphanumeric characters with underscores
      str_to_lower()  # Convert to lowercase
  )

# 4. **Pivot to match new and old files**
# Reshape so each row = one logical file, with separate columns for old/new full paths.
# Files present in only one version will have NA in the other path column.
file_level_log <- all_files %>%
  pivot_wider(
    id_cols = files_std,             # Only use files_std to identify matching files
    names_from = directory, 
    values_from = full_path, 
    names_prefix = ""
  ) %>%
  mutate(
    exists_in_old  = !is.na(old),                                            # TRUE if file present in old version
    exists_in_new  = !is.na(new),                                            # TRUE if file present in new version
    size_old_mb    = ifelse(!is.na(old), file.info(old)$size / 1e6, NA_real_), # File size in MB for old version
    size_new_mb    = ifelse(!is.na(new), file.info(new)$size / 1e6, NA_real_), # File size in MB for new version
    size_change_mb = size_new_mb - size_old_mb                               # Positive = file grew, negative = file shrunk
  )

# 5. **Data-Level Change Log**
# Restrict comparisons to files that exist in both versions and are the same
# format (.csv-.csv or .xlsx-.xlsx); PDFs and unmatched files are excluded
comparison_pairs <- file_level_log %>%
  filter(exists_in_old & exists_in_new) %>%
  # compare matching .csv files 
  filter((str_detect(new, "\\.csv$") & str_detect(old, "\\.csv$")) |
           # OR matching .xlsx files
           (str_detect(new, "\\.xlsx$") & str_detect(old, "\\.xlsx$"))) %>%
  select(files_std, new, old)


# Read a pair of files and return a named list of comparison metrics:
# row counts, column names, added/removed columns, and per-column data type changes
compare_files <- function(file_new, file_old) {
  if(basename(file_new) %>% str_detect("\\.csv$")) {
    data_new <- fread(file_new)
  } 
  if(basename(file_new) %>% str_detect("\\.xlsx$")) {
    data_new <- readxl::read_xlsx(file_new, skip=1)
  }
  if(basename(file_old) %>% str_detect("\\.csv$")) {
    data_old <- fread(file_old)
  }
  if(basename(file_old) %>% str_detect("\\.xlsx$")) {
    data_old <- readxl::read_xlsx(file_old, skip=1)
  }
  
  list(
    file_name_new = basename(file_new),
    file_name_old = basename(file_old),
    new_rows      = nrow(data_new),
    old_rows      = nrow(data_old),
    row_diff      = nrow(data_new) - nrow(data_old),  # Positive = rows added, negative = rows removed
    new_cols      = colnames(data_new),
    old_cols      = colnames(data_old),
    # single col value will create a vector column - >1 cols will create list-col
    added_cols = if (length(setdiff(colnames(data_new),
                                    colnames(data_old))) > 0) {
      setdiff(colnames(data_new), colnames(data_old))   # Columns in new but not old
    } else { NA_character_},
    removed_cols = if (length(setdiff(colnames(data_old),
                                      colnames(data_new))) > 0) {
      setdiff(colnames(data_old), colnames(data_new))   # Columns in old but not new
    } else {NA_character_},
                   
    # Build a per-column type comparison across the union of all column names
    data_type_changes = tibble(
      column = union(colnames(data_new), colnames(data_old)),
      # if col is in data_old get its class - else it is NA_character_
      old_type = map_chr(union(colnames(data_new), colnames(data_old)), 
                         ~ ifelse(.x %in% colnames(data_old), 
                                  class(data_old[[.x]])[1], 
                                  NA_character_)),
      # if col is in data_new get its class - else it is NA_character_
      new_type = map_chr(union(colnames(data_new), 
                               colnames(data_old)), 
                         ~ ifelse(.x %in% colnames(data_new), 
                                  class(data_new[[.x]])[1], 
                                  NA_character_))
    ) %>% 
      mutate(type_changed = old_type != new_type)  # Flag columns where the type differs
  )
}

# Apply compare_files() across all matched pairs; result is a list-column
data_level_log <- comparison_pairs %>%
  mutate(comparison = map2(new, old, ~ compare_files(.x, .y)))

# Extract data-level comparison details
# Unnest the list-column into individual columns for downstream use
data_change_log <- data_level_log %>%
  unnest_wider(comparison) %>%
  select(file_name_new, file_name_old, new_rows, old_rows, row_diff, 
         added_cols, removed_cols, data_type_changes) 

```

```{r fao-file-table}
#| include: TRUE

# Re-derive files_std from data_change_log file names (same logic as fao-scan-files)
# so we can join row counts back into the file-level log for display
file_table <- file_level_log |>
  left_join(
    data_change_log |>
      mutate(files_std = file_name_new |>
               str_remove("_V\\d{6,7}[a-z]*") |>
               str_replace_all("\\.[a-zA-Z0-9]+$", "") |>
               str_replace_all("[^a-zA-Z0-9]", "_") |>
               str_to_lower()
      ) |>
      select(files_std, new_rows, old_rows, row_diff),
    by = "files_std"
  ) |>
  mutate(
    file_name     = coalesce(basename(new), basename(old)),  # Use new name if available, fall back to old
    exists_in_old = if_else(exists_in_old, "✓", "—"),
    exists_in_new = if_else(exists_in_new, "✓", "—"),
    .flag_new     = exists_in_old == "—",   # new file (not in old)
    .flag_gone    = exists_in_new == "—"    # removed file (not in new)
  ) |>
  select(
    file_name,
    exists_in_old, exists_in_new,
    size_old_mb, size_new_mb, size_change_mb,
    old_rows, new_rows, row_diff,
    .flag_new, .flag_gone  # retained for row highlighting below; dropped before rendering
  )

file_table |>
  select(-.flag_new, -.flag_gone) |>  # Drop flag columns — used for styling only
  gt() |>
  tab_header(
    title    = "FAO Global Production — File-Level Assessment",
    subtitle = glue("{fao_v_old} vs {fao_v_new}")
  ) |>
  cols_label(
    file_name      = "File",
    exists_in_old  = glue("In {fao_v_old}"),
    exists_in_new  = glue("In {fao_v_new}"),
    size_old_mb    = glue("Size {fao_v_old} (MB)"),
    size_new_mb    = glue("Size {fao_v_new} (MB)"),
    size_change_mb = "Size Δ (MB)",
    old_rows       = glue("Rows {fao_v_old}"),
    new_rows       = glue("Rows {fao_v_new}"),
    row_diff       = "Row Δ"
  ) |>
  fmt_number(
    columns  = c(size_old_mb, size_new_mb, size_change_mb),
    decimals = 4
  ) |>
  fmt_integer(
    columns  = c(old_rows, new_rows, row_diff),
    use_seps = TRUE
  ) |>
  sub_missing(missing_text = "—") |>
  # Highlight new files (green) and removed files (red)
  tab_style(
    style     = cell_fill(color = "#d9ead3"),
    locations = cells_body(rows = file_table$.flag_new)
  ) |>
  tab_style(
    style     = cell_fill(color = "#f4cccc"),
    locations = cells_body(rows = file_table$.flag_gone)
  ) |>
  # Bold the row diff value when it is non-zero to draw attention to row count changes
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_body(
      columns = row_diff,
      rows    = !is.na(file_table$row_diff) & file_table$row_diff != 0
    )
  ) |>
  tab_footnote(
    footnote  = "Green rows = files new in the incoming version. Red rows = files removed from the incoming version.",
    locations = cells_column_labels(columns = exists_in_new)
  ) |>
  tab_options(
    table.font.size           = 12,
    column_labels.font.weight = "bold",
    heading.align             = "left"
  )
```

## Production Quantity

::: {.callout-note}
Not sure what we expect between versions, consistent quantity? OR is some difference ok? Created row highlighting based on percent difference threshold value at the top. 
:::

```{r fao-production-quantity}
#| include: TRUE

# Read the raw quantity CSVs directly — no rebuild_fao_2023_dat() involved.
# Raw column names from the FAO format: PERIOD (year), VALUE (production quantity).
qty_raw_old <- fread(fao_gp_old)
qty_raw_new <- fread(fao_gp_new)

# Derive year range from raw data (replaces fao-year-coverage chunk)
min_yr_old <- min(qty_raw_old$PERIOD, na.rm = TRUE)
max_yr_old <- max(qty_raw_old$PERIOD, na.rm = TRUE)
min_yr_new <- min(qty_raw_new$PERIOD, na.rm = TRUE)
max_yr_new <- max(qty_raw_new$PERIOD, na.rm = TRUE)

# Summarise total production quantity per year, filtered to 1996+
qty_old <- qty_raw_old |>
  filter(PERIOD >= 1996) |>
  group_by(year = PERIOD) |>
  summarise(total_qty_old = sum(VALUE, na.rm = TRUE), .groups = "drop")

qty_new <- qty_raw_new |>
  filter(PERIOD >= 1996) |>
  group_by(year = PERIOD) |>
  summarise(total_qty_new = sum(VALUE, na.rm = TRUE), .groups = "drop")

# Full join so years present in only one version are retained
qty_compare <- full_join(qty_old, qty_new, by = "year") |>
  mutate(
    qty_diff      = total_qty_new - total_qty_old,
    qty_diff_pct  = round(100 * qty_diff / total_qty_old, 2),
    # Row highlight flags — not rendered in the table
    .flag_old_only  = !is.na(total_qty_old) & is.na(total_qty_new),                        # year in old, missing in new
    .flag_new_only  = is.na(total_qty_old) & !is.na(total_qty_new),                        # year in new only (expected new years)
    .flag_threshold = !is.na(qty_diff_pct) & abs(qty_diff_pct) >= qty_pct_threshold         # % change exceeds threshold
  ) |>
  arrange(year)

qty_compare |>
  select(-.flag_old_only, -.flag_new_only, -.flag_threshold) |>  # Drop flag columns — used for styling only
  gt() |>
  tab_header(
    title    = "FAO Global Production — Annual Quantity Assessment",
    subtitle = glue("{fao_v_old} vs {fao_v_new}")
  ) |>
  cols_label(
    year          = "Year",
    total_qty_old = glue("Total Qty ({fao_v_old})"),
    total_qty_new = glue("Total Qty ({fao_v_new})"),
    qty_diff      = "Qty Δ",
    qty_diff_pct  = "Qty Δ (%)"
  ) |>
  fmt_integer(
    columns  = c(total_qty_old, total_qty_new, qty_diff),
    use_seps = TRUE
  ) |>
  fmt_number(
    columns  = qty_diff_pct,
    decimals = 2
  ) |>
  sub_missing(missing_text = "—") |>
  # Red: years present in old but missing in new (unexpected)
  tab_style(
    style     = cell_fill(color = "#f4cccc"),
    locations = cells_body(rows = qty_compare$.flag_old_only)
  ) |>
  # Neutral yellow: years present in new only (expected new data years)
  tab_style(
    style     = cell_fill(color = "#fff2cc"),
    locations = cells_body(rows = qty_compare$.flag_new_only)
  ) |>
  # Purple: absolute % change exceeds threshold — warrants review
  tab_style(
    style     = cell_fill(color = "#d9d2e9"),
    locations = cells_body(rows = qty_compare$.flag_threshold)
  ) |>
  tab_footnote(
    footnote  = glue("Purple rows = absolute % change equals or exceeds {qty_pct_threshold}% threshold and warrants review."),
    locations = cells_column_labels(columns = qty_diff_pct)
  ) |>
  tab_footnote(
    footnote  = "Yellow rows = years new in the incoming version. Red rows = years present in the previous version but missing from the incoming version.",
    locations = cells_column_labels(columns = year)
  ) |>
  tab_options(
    table.font.size           = 12,
    column_labels.font.weight = "bold",
    heading.align             = "left"
  )
```

# Assess: FAO Global Production Taxa

## Taxa unique to new ``r fao_v_new`` version

```{r fao-sciname-unique-new}
#| include: TRUE
#| results: 'asis'

# get old production species - uncleaned
raw_prod_old_fn <- glue("GlobalProduction_{fao_v_old}.zip")
raw_prod_old_sp <- artis::rebuild_fao_2023_dat(
  datadir = path_raw_data_old, 
  filename = raw_prod_old_fn) %>%
  select(
    sciname = species_scientific_name, 
    common_name = species_name_en) %>%
  mutate(
    sciname = tolower(sciname),
    common_name = tolower(common_name)) %>%
  distinct()

# get new production species - uncleaned
raw_prod_new_fn <- glue("GlobalProduction_{fao_v_new}.zip")
raw_prod_new_sp <- artis::rebuild_fao_2023_dat(
  datadir = path_raw_data_new, 
  filename = raw_prod_new_fn) %>%
  select(
    sciname = species_scientific_name, 
    common_name = species_name_en) %>%
  mutate(
    sciname = tolower(sciname),
    common_name = tolower(common_name)) %>%
  distinct()

raw_prod_old_sp_only <- anti_join(raw_prod_old_sp, raw_prod_new_sp, by = "sciname") %>%
  arrange(sciname) %>% 
  mutate(version = fao_v_old)

raw_prod_new_sp_only <- anti_join(raw_prod_new_sp, raw_prod_old_sp, by = "sciname") %>%
  arrange(sciname)  %>% 
  mutate(version = fao_v_new)

# --- Table 1: Taxa unique to new version ---
new_sp_tbl <- raw_prod_new_sp_only |>
  mutate(likely_hybrid = str_detect(sciname, "\\bx\\b"))

new_sp_tbl |>
  select(sciname, common_name, version, likely_hybrid) |>
  gt() |>
  tab_header(
    title    = glue("FAO Global Production — New Taxa Assessment"),
    subtitle = glue("{fao_v_old} vs {fao_v_new}")
  ) |>
  cols_label(
    sciname       = "Scientific Name",
    common_name   = "Common Name",
    version       = "Version",
    likely_hybrid = "Hybrid"
  ) |>
  fmt(
    columns = likely_hybrid,
    fns     = function(x) ifelse(x, "✓", "—")
  ) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_fill(color = "#fff2cc"),
    locations = cells_body(rows = new_sp_tbl$likely_hybrid)
  ) |>
  tab_options(
    table.font.size           = 12,
    column_labels.font.weight = "bold",
    heading.align             = "left",
    row.striping.include_table_body = TRUE
  )
```

## Taxa unique to old ``r fao_v_old`` version

```{r fao-unique-taxa-old}
#| include: TRUE
#| results: 'asis'
# --- Table 2: Taxa unique to old version ---
old_sp_tbl <- raw_prod_old_sp_only |>
  mutate(likely_hybrid = str_detect(sciname, "\\bx\\b"))

old_sp_tbl |>
  select(sciname, common_name, version, likely_hybrid) |>
  gt() |>
  tab_header(
    title    = glue("FAO Global Production — Removed Taxa Assessment"),
    subtitle = glue("{fao_v_old} vs {fao_v_new}")
  ) |>
  cols_label(
    sciname       = "Scientific Name",
    common_name   = "Common Name",
    version       = "Version",
    likely_hybrid = "Hybrid"
  ) |>
  fmt(
    columns = likely_hybrid,
    fns     = function(x) ifelse(x, "✓", "—")
  ) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_fill(color = "#fff2cc"),
    locations = cells_body(rows = old_sp_tbl$likely_hybrid)
  ) |>
  tab_options(
    table.font.size           = 12,
    column_labels.font.weight = "bold",
    heading.align             = "left",
    row.striping.include_table_body = TRUE
  )
```

```{r fao-common-names}
#| include: TRUE
#| results: 'asis'
common_name_changes <- bind_rows(
  anti_join(
    raw_prod_old_sp |> distinct(sciname, common_name),
    raw_prod_new_sp |> distinct(sciname, common_name),
    by = c("sciname", "common_name")
  ) |> mutate(change = "removed"),
  anti_join(
    raw_prod_new_sp |> distinct(sciname, common_name),
    raw_prod_old_sp |> distinct(sciname, common_name),
    by = c("sciname", "common_name")
  ) |> mutate(change = "added")
) |>
  arrange(sciname, change)

shared_scinames <- intersect(
  raw_prod_old_sp$sciname,
  raw_prod_new_sp$sciname
)

common_name_changes_filtered <- common_name_changes |>
  filter(sciname %in% shared_scinames)

# derive alternating group index per sciname
sciname_groups <- common_name_changes_filtered |>
  distinct(sciname) |>
  mutate(.grp = row_number() %% 2 == 0)  # TRUE = even rows get shading

common_name_changes_filtered |>
  left_join(sciname_groups, by = "sciname") |>
  gt() |>
  tab_header(
    title    = "FAO Global Production — Common Name Changes",
    subtitle = glue("{fao_v_old} vs {fao_v_new}")
  ) |>
  cols_label(
    sciname     = "Scientific Name",
    common_name = "Common Name",
    change      = "Change"
  ) |>
  cols_hide(.grp) |>
  tab_style(
    style     = cell_fill(color = "#efefef"),
    locations = cells_body(
      rows = common_name_changes_filtered |>
        left_join(sciname_groups, by = "sciname") |>
        pull(.grp)
    )
  ) |>
  tab_style(
    style     = cell_text(color = "#2a7a2a", weight = "bold"),
    locations = cells_body(columns = change, rows = common_name_changes_filtered$change == "added")
  ) |>
  tab_style(
    style     = cell_text(color = "#b94040", weight = "bold"),
    locations = cells_body(columns = change, rows = common_name_changes_filtered$change == "removed")
  ) |>
  tab_options(
    table.font.size           = 12,
    column_labels.font.weight = "bold",
    heading.align             = "left"
  )
```

# Assess: Fishbase / Sealifebase Taxa

```{r fbslb-taxa-changes}
#| include: TRUE
#| results: 'asis'
path_fbslb_new <- file.path(path_raw_data_new, glue("fishbase_sealifebase_{fbslb_v_new}"))
path_fbslb_old <- file.path(path_raw_data_old, glue("fishbase_sealifebase_{fbslb_v_old}"))

# Load taxa info for both databases and both versions
fb_taxa_old  <- fread(file.path(path_fbslb_old, "fb_taxa_info.csv"))
fb_taxa_new  <- fread(file.path(path_fbslb_new, "fb_taxa_info.csv"))
slb_taxa_old <- fread(file.path(path_fbslb_old, "slb_taxa_info.csv"))
slb_taxa_new <- fread(file.path(path_fbslb_new, "slb_taxa_info.csv"))

# Tag each with its database label before combining
fb_taxa_old  <- fb_taxa_old  |> mutate(database = "FishBase")
fb_taxa_new  <- fb_taxa_new  |> mutate(database = "FishBase")
slb_taxa_old <- slb_taxa_old |> mutate(database = "SeaLifeBase")
slb_taxa_new <- slb_taxa_new |> mutate(database = "SeaLifeBase")

taxa_old <- bind_rows(fb_taxa_old, slb_taxa_old)
taxa_new <- bind_rows(fb_taxa_new, slb_taxa_new)

display_cols <- c("database", "SpecCode", "Species", "Genus", "Family", "Order", "Class")

# Taxa added in new version (present in new, absent in old) — per database
taxa_added <- bind_rows(
  anti_join(fb_taxa_new,  fb_taxa_old,  by = c("database", "SpecCode")),
  anti_join(slb_taxa_new, slb_taxa_old, by = c("database", "SpecCode"))
) |>
  select(any_of(display_cols)) |>
  mutate(change = "added")

# Taxa removed in new version (present in old, absent in new) — per database
taxa_removed <- bind_rows(
  anti_join(fb_taxa_old,  fb_taxa_new,  by = c("database", "SpecCode")),
  anti_join(slb_taxa_old, slb_taxa_new, by = c("database", "SpecCode"))
) |>
  select(any_of(display_cols)) |>
  mutate(change = "removed")

taxa_changes <- bind_rows(taxa_added, taxa_removed) |>
  arrange(database, change, Species)
```

```{r fb-taxa-added-tbl}
#| include: TRUE

taxa_fb_added <- taxa_changes |>
  filter(change == "added", database == "FishBase") |>
  arrange(Species)

taxa_fb_added |>
  select(Species, Genus, Family, Order, Class) |>
  gt() |>
  tab_header(
    title    = "FishBase — Taxa Added",
    subtitle = glue("{fbslb_v_old} vs {fbslb_v_new}")
  ) |>
  cols_label(
    Species = "Scientific Name",
    Genus   = "Genus",
    Family  = "Family",
    Order   = "Order",
    Class   = "Class"
  ) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_text(color = "#2a7a2a", weight = "bold"),
    locations = cells_body(columns = Species)
  ) |>
  tab_options(
    table.font.size                 = 11,
    column_labels.font.weight      = "bold",
    heading.align                   = "left",
    row.striping.include_table_body = TRUE,
    row.striping.background_color   = "#efefef"
  ) |>
  opt_interactive(
    use_search           = TRUE,
    use_pagination       = FALSE,
    use_page_size_select = FALSE
  )
```

```{r slb-taxa-added-tbl}
#| include: TRUE

taxa_slb_added <- taxa_changes |>
  filter(change == "added", database == "SeaLifeBase") |>
  arrange(Species)

taxa_slb_added |>
  select(Species, Genus, Family, Order, Class) |>
  gt() |>
  tab_header(
    title    = "SeaLifeBase — Taxa Added",
    subtitle = glue("{fbslb_v_old} vs {fbslb_v_new}")
  ) |>
  cols_label(
    Species = "Scientific Name",
    Genus   = "Genus",
    Family  = "Family",
    Order   = "Order",
    Class   = "Class"
  ) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_text(color = "#2a7a2a", weight = "bold"),
    locations = cells_body(columns = Species)
  ) |>
  tab_options(
    table.font.size                 = 11,
    column_labels.font.weight      = "bold",
    heading.align                   = "left",
    row.striping.include_table_body = TRUE,
    row.striping.background_color   = "#efefef"
  ) |>
  opt_interactive(
    use_search           = TRUE,
    use_pagination       = FALSE,
    use_page_size_select = FALSE
  )
```

```{r fb-taxa-removed-tbl}
#| include: TRUE

taxa_fb_removed <- taxa_changes |>
  filter(change == "removed", database == "FishBase") |>
  arrange(Species)

taxa_fb_removed |>
  select(Species, Genus, Family, Order, Class) |>
  gt() |>
  tab_header(
    title    = "FishBase — Taxa Removed",
    subtitle = glue("{fbslb_v_old} vs {fbslb_v_new}")
  ) |>
  cols_label(
    Species = "Scientific Name",
    Genus   = "Genus",
    Family  = "Family",
    Order   = "Order",
    Class   = "Class"
  ) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_text(color = "#b94040", weight = "bold"),
    locations = cells_body(columns = Species)
  ) |>
  tab_options(
    table.font.size                 = 11,
    column_labels.font.weight      = "bold",
    heading.align                   = "left",
    row.striping.include_table_body = TRUE,
    row.striping.background_color   = "#efefef"
  ) |>
  opt_interactive(
    use_search           = TRUE,
    use_pagination       = FALSE,
    use_page_size_select = FALSE
  )
```

## NOT Validated yet: 

```{r slb-taxa-removed-tbl}
#| include: FALSE
#| eval: false

taxa_slb_removed <- taxa_changes |>
  filter(change == "removed", database == "SeaLifeBase") |>
  arrange(Species)

taxa_slb_removed |>
  select(Species, Genus, Family, Order, Class) |>
  gt() |>
  tab_header(
    title    = "SeaLifeBase — Taxa Removed",
    subtitle = glue("{fbslb_v_old} vs {fbslb_v_new}")
  ) |>
  cols_label(
    Species = "Scientific Name",
    Genus   = "Genus",
    Family  = "Family",
    Order   = "Order",
    Class   = "Class"
  ) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_text(color = "#b94040", weight = "bold"),
    locations = cells_body(columns = Species)
  ) |>
  tab_options(
    table.font.size                 = 11,
    column_labels.font.weight      = "bold",
    heading.align                   = "left",
    row.striping.include_table_body = TRUE,
    row.striping.background_color   = "#efefef"
  ) |>
  opt_interactive(
    use_search           = TRUE,
    use_pagination       = FALSE,
    use_page_size_select = FALSE
  )
```

```{r fbslb-taxonomy-diff}
#| include: FALSE
#| results: 'asis'
#| eval: false

tax_ranks <- c("Genus", "Family", "Order", "Class")

# For species present in BOTH versions, pivot to long and detect rank-level changes.
# Done separately per database so SpecCode is unambiguous, then combined.
make_tax_diff <- function(old_df, new_df, db_label) {
  shared_codes <- intersect(old_df$SpecCode, new_df$SpecCode)

  old_long <- old_df |>
    filter(SpecCode %in% shared_codes) |>
    select(SpecCode, Species, all_of(tax_ranks)) |>
    pivot_longer(all_of(tax_ranks), names_to = "rank", values_to = "old_value")

  new_long <- new_df |>
    filter(SpecCode %in% shared_codes) |>
    select(SpecCode, Species, all_of(tax_ranks)) |>
    pivot_longer(all_of(tax_ranks), names_to = "rank", values_to = "new_value")

  old_long |>
    left_join(new_long, by = c("SpecCode", "Species", "rank")) |>
    filter(old_value != new_value | (is.na(old_value) != is.na(new_value))) |>
    mutate(
      database  = db_label,
      rank      = factor(rank, levels = tax_ranks)
    ) |>
    select(database, SpecCode, Species, rank, old_value, new_value)
}

tax_diff <- bind_rows(
  make_tax_diff(fb_taxa_old,  fb_taxa_new,  "FishBase"),
  make_tax_diff(slb_taxa_old, slb_taxa_new, "SeaLifeBase")
) |>
  arrange(database, Species, rank)
```




```{r fbslb-taxonomy-diff-tbl}
#| include: FALSE
#| results: 'asis'
#| eval: false

if (nrow(tax_diff) == 0) {
  cat("> **No higher-level taxonomy changes detected between versions.**")
} else {
  # Alternating stripe by Species (same pattern as fao-common-names)
  tax_diff_groups <- tax_diff |>
    distinct(database, Species) |>
    arrange(database, Species) |>
    mutate(.grp = row_number() %% 2 == 0)

  tax_diff_display <- tax_diff |>
    left_join(tax_diff_groups, by = c("database", "Species"))

  tax_diff_display |>
    select(database, Species, rank, old_value, new_value, .grp) |>
    gt() |>
    tab_header(
      title    = "FishBase / SeaLifeBase — Higher-Level Taxonomy Changes",
      subtitle = glue("{fbslb_v_old} vs {fbslb_v_new}")
    ) |>
    cols_label(
      database  = "Database",
      Species   = "Scientific Name",
      rank      = "Rank",
      old_value = glue("Old ({fbslb_v_old})"),
      new_value = glue("New ({fbslb_v_new})")
    ) |>
    cols_hide(.grp) |>
    sub_missing(missing_text = "—") |>
    tab_style(
      style     = cell_fill(color = "#efefef"),
      locations = cells_body(rows = tax_diff_display$.grp)
    ) |>
    tab_style(
      style     = cell_text(color = "#b94040", weight = "bold"),
      locations = cells_body(columns = old_value)
    ) |>
    tab_style(
      style     = cell_text(color = "#2a7a2a", weight = "bold"),
      locations = cells_body(columns = new_value)
    ) |>
    tab_options(
      table.font.size           = 11,
      column_labels.font.weight = "bold",
      heading.align             = "left"
    )
}
```

```{r fbslb-synonym-changes}
#| include: FALSE
#| results: 'asis'
#| eval: false

fb_syn_old  <- fread(file.path(path_fbslb_old, "fb_synonyms_clean.csv"))
fb_syn_new  <- fread(file.path(path_fbslb_new, "fb_synonyms_clean.csv"))
slb_syn_old <- fread(file.path(path_fbslb_old, "slb_synonyms_clean.csv"))
slb_syn_new <- fread(file.path(path_fbslb_new, "slb_synonyms_clean.csv"))

# Key columns from clean_fb_slb_synonyms(): synonym, accepted_name, spec_code
syn_key <- c("synonym", "accepted_name", "spec_code")

make_syn_diff <- function(old_df, new_df, db_label) {
  bind_rows(
    anti_join(old_df, new_df, by = syn_key) |> mutate(change = "removed"),
    anti_join(new_df, old_df, by = syn_key) |> mutate(change = "added")
  ) |>
    mutate(database = db_label) |>
    select(database, synonym, accepted_name, change)
}

syn_changes <- bind_rows(
  make_syn_diff(fb_syn_old,  fb_syn_new,  "FishBase"),
  make_syn_diff(slb_syn_old, slb_syn_new, "SeaLifeBase")
) |>
  arrange(database, synonym, change)
```

```{r fbslb-synonym-changes-tbl}
#| include: FALSE
#| results: 'asis'
#| eval: false

if (nrow(syn_changes) == 0) {
  cat("> **No synonym mapping changes detected between versions.**")
} else {
  syn_groups <- syn_changes |>
    distinct(database, synonym) |>
    arrange(database, synonym) |>
    mutate(.grp = row_number() %% 2 == 0)

  syn_display <- syn_changes |>
    left_join(syn_groups, by = c("database", "synonym"))

  syn_display |>
    select(database, synonym, accepted_name, change, .grp) |>
    gt() |>
    tab_header(
      title    = "FishBase / SeaLifeBase — Synonym Mapping Changes",
      subtitle = glue("{fbslb_v_old} vs {fbslb_v_new}")
    ) |>
    cols_label(
      database      = "Database",
      synonym       = "Synonym",
      accepted_name = "Accepted Name",
      change        = "Change"
    ) |>
    cols_hide(.grp) |>
    sub_missing(missing_text = "—") |>
    tab_style(
      style     = cell_fill(color = "#efefef"),
      locations = cells_body(rows = syn_display$.grp)
    ) |>
    tab_style(
      style     = cell_fill(color = "#fff2cc"),
      locations = cells_body(rows = seq_len(nrow(syn_display)))
    ) |>
    tab_options(
      table.font.size           = 11,
      column_labels.font.weight = "bold",
      heading.align             = "left"
    )
}
```

# Assess: BACI International Bilateral Trade

```{r}
path_baci_old <- file.path(path_raw_data_old, "baci_raw")
path_baci_new <- file.path(path_raw_data_new, "baci_raw")
```

## Countries

```{r}
#| eval: false
# Old BACI assessment ----------------------------------

# Get HS version from raw baci data directories
hs_versions_old <- list.files(
  path = path_baci_old,
  pattern = "HS\\d{2}",
  recursive = FALSE
) |>
  str_subset("\\.zip$", negate = TRUE) |>
  str_extract("HS\\d{2}")

countries_old <- list.files(
  path = path_baci_old,
  pattern = "country_codes",
  recursive = TRUE
) 
# Check assumption that all country_code_<baci_version..csv files are identical
countries_old_data <- file.path(path_baci_old, countries_old) |>
  map(fread)

all_identical <- countries_old_data |>
  map_lgl(\(x) identical(as.data.frame(x), as.data.frame(countries_old_data[[1]]))) |>
  all()

all_identical

# log in the validation log
validation_log[[glue("BACI {baci_v_old} country code files identical across HS versions")]] <- list(
  passed = all_identical,
  notes  = "Check for differing file versions in path_baci_old. Files should all share the same V-version suffix."
)

# New Baci assessment ------------------------------
# Get HS version from raw baci data directories
hs_versions_new <- list.files(
  path = path_baci_new,
  pattern = "HS\\d{2}",
  recursive = FALSE
) |>
  str_subset("\\.zip$", negate = TRUE) |>
  str_extract("HS\\d{2}")

countries_new <- list.files(
  path = path_baci_new,
  pattern = "country_codes",
  recursive = TRUE
) 
# Check assumption that all country_code_<baci_version..csv files are identical
countries_new_data <- file.path(path_baci_new, countries_new) |>
  map(fread)

all_identical <- countries_new_data |>
  map_lgl(\(x) identical(as.data.frame(x), as.data.frame(countries_new_data[[1]]))) |>
  all()

all_identical

# log in the validation log
validation_log[[glue("BACI {baci_v_new} country code files identical across HS versions")]] <- list(
  passed = all_identical,
  notes  = "Check for differing file versions in path_baci_new. Files should all share the same V-version suffix."
)
```

## Products
```{r}
#| eval: FALSE


# Get all product_codes files for old and new BACI
products_old <- list.files(
  path = path_baci_old,
  pattern = "product_codes",
  recursive = TRUE
)

products_new <- list.files(
  path = path_baci_new,
  pattern = "product_codes",
  recursive = TRUE
)

# Build tidy tibbles with HS version extracted, then join to pair by HS version
products_old_df <- tibble(rel_path = products_old) |>
  mutate(
    hs_version  = str_extract(rel_path, "HS\\d{2}"),
    full_path   = file.path(path_baci_old, rel_path),
    baci_v      = baci_v_old
  )

products_new_df <- tibble(rel_path = products_new) |>
  mutate(
    hs_version  = str_extract(rel_path, "HS\\d{2}"),
    full_path   = file.path(path_baci_new, rel_path),
    baci_v      = baci_v_new
  )

# Pair old and new by HS version
products_paired <- inner_join(
  products_old_df, 
  products_new_df,
  by = join_by(hs_version),
  suffix = c("_old", "_new")
)

# Compare each HS version pair
compare_product_codes <- function(path_old, path_new, hs_ver) {
  old_dat <- fread(path_old)
  new_dat <- fread(path_new)

  codes_removed <- anti_join(
    as.data.frame(old_dat), as.data.frame(new_dat), by = "code"
  )
  codes_added <- anti_join(
    as.data.frame(new_dat), as.data.frame(old_dat), by = "code"
  )

  tibble(
    hs_version      = hs_ver,
    n_codes_old     = nrow(old_dat),
    n_codes_new     = nrow(new_dat),
    n_codes_removed = nrow(codes_removed),
    n_codes_added   = nrow(codes_added),
    codes_removed   = list(codes_removed),
    codes_added     = list(codes_added)
  )
}

product_code_comparison <- products_paired |>
  pmap(
    \(hs_version, full_path_old, full_path_new, ...) 
      compare_product_codes(full_path_old, full_path_new, hs_version)
  ) |>
  list_rbind()

# Log validation checks per HS version
walk(product_code_comparison$hs_version, \(hs_ver) {
  row <- filter(product_code_comparison, hs_version == hs_ver)
  passed <- row$n_codes_removed == 0 & row$n_codes_added == 0
  validation_log[[glue("BACI {hs_ver} product codes unchanged {baci_v_old} -> {baci_v_new}")]] <<- list(
    passed = passed,
    notes  = glue("Check product_codes_{hs_ver}_*.csv files in path_baci_old and path_baci_new. {row$n_codes_added} codes added, {row$n_codes_removed} codes removed.")
  )
})

# Display product code comparison summary table
product_code_comparison |>
  select(hs_version, n_codes_old, n_codes_new, n_codes_removed, n_codes_added) |>
  gt() |>
  tab_header(title = "BACI Product Code Changes by HS Version",
             subtitle = glue("{baci_v_old} vs {baci_v_new}")) |>
  cols_label(
    hs_version      = "HS Version",
    n_codes_old     = glue("# Codes ({baci_v_old})"),
    n_codes_new     = glue("# Codes ({baci_v_new})"),
    n_codes_removed = "# Removed",
    n_codes_added   = "# Added"
  ) |>
  tab_style(
    style = cell_fill(color = "#f4cccc"),
    locations = cells_body(rows = n_codes_removed > 0 | n_codes_added > 0)
  )
```
