---
title: "ARTIS Data Ingest"
subtitle: "Changelog of raw data verions"
author: "Althea Marks"
format:
  pdf:
    papersize: letter
execute:
  echo: FALSE
  include: FALSE
  message: FALSE
  warning: FALSE
---

```{r setup}

#### Library #####
{
  library(data.table)
  library(dplyr)
  library(readr)
  library(stringr)
  library(purrr)
  library(tidyr)
  library(knitr)
  library(kableExtra)
  library(readxl)
  library(gt)
  library(artis)
  library(glue)
  library(testthat)
  library(readxl)
}


#### Directory setup #####

# FIXIT - use setup script to streamline
#source("00-local-machine-setup.R")

# Newest data
datadir_new <- file.path("~", "Documents", "UW-SAFS", "ARTIS", "data",
                         "model_inputs_raw_3.0")
# last data ingest
datadir_old <- file.path("~", "Documents", "UW-SAFS", "ARTIS", "data",
                         "model_inputs_raw_2.0")
modelfundir <- file.path("~", "Documents", "git-projects", "artis-model", "R")


# values for report
raw_data_type <- "FAO Global Production"
artis_old <- "2.0"
artis_new <- "3.0"
data_v_old <- "2025.1.0"
data_v_new <- "2026.1.0"
baci_v_old <- "V202501"
baci_v_new <- "V202601"

# create validation log empty list
validation_log <- list()
```

```{r get-prod-files}

# List all subdirectories in datadir_new
subdirs_new <- list.dirs(datadir_new, recursive = TRUE, full.names = TRUE)
subdirs_old <- list.dirs(datadir_old, recursive = TRUE, full.names = TRUE)

# FAO Global production regex pattern
pattern_fao <- regex("(global[\\s_-]*production)", ignore_case = TRUE)

# FAO Global production directory
fao_dir_new <- subdirs_new[str_detect(subdirs_new, pattern_fao)]
fao_dir_old <- subdirs_old[str_detect(subdirs_old, pattern_fao)]

gp_new_fp <- file.path(fao_dir_new, "Global_production_quantity.csv")
gp_old_fp <- file.path(fao_dir_old, "Global_production_quantity.csv")

# BACI regex pattern
# pattern_baci <- regex("(baci)", ignore_case = TRUE)
# 
# # BACI directory
# baci_dir_new <- subdirs_new[str_detect(subdirs_new, pattern_baci)]
#baci_dir_old <- subdirs_old[str_detect(subdirs_old, pattern_baci)]

```

```{r assess-prod-file-changes}

# 1. **Scan directories**
scan_dir <- function(dir, rg = "") {
  if (length(dir) == 0) return(character(0)) # Handle empty directories
  list.files(dir, pattern = rg, recursive = TRUE, full.names = TRUE)
}

# 2. **Read files from directories**
files_new <- scan_dir(fao_dir_new)
files_old <- scan_dir(fao_dir_old)

# 3. **Standardize file names**
all_files <- tibble(
  full_path = c(files_new, files_old),
  files = c(basename(files_new), basename(files_old)),
  directory = c(rep("new", length(files_new)), rep("old", length(files_old)))
) %>%
  mutate(
    files_std = files %>% 
      str_remove("^filtered_") %>%  # Remove "filtered_" prefix
      str_remove("_V\\d{6,7}[a-z]*") %>% # Remove version suffix (like _V202201)
      str_replace_all("\\.[a-zA-Z0-9]+$", "") %>%  # Remove extensions like .csv
      str_replace_all("[^a-zA-Z0-9]", "_") %>%   # Replace non-alphanumeric characters with underscores
      str_to_lower()  # Convert to lowercase
  )

# 4. **Pivot to match new and old files**
file_level_log <- all_files %>%
  pivot_wider(
    id_cols = files_std,             # Only use files_std to identify matching files
    names_from = directory, 
    values_from = full_path, 
    names_prefix = ""
  ) %>%
  mutate(
    exists_in_old = !is.na(old),
    exists_in_new = !is.na(new),
    size_old_mb = ifelse(!is.na(old), 
                         file.info(old)$size / 1e6, 
                         NA_real_),
    size_new_mb = ifelse(!is.na(new), 
                         file.info(new)$size / 1e6, 
                         NA_real_),
    size_change_mb = size_new_mb - size_old_mb
  )

# 5. **Data-Level Change Log**
comparison_pairs <- file_level_log %>%
  filter(exists_in_old & exists_in_new) %>%
  # compare matching .csv files 
  filter((str_detect(new, "\\.csv$") & str_detect(old, "\\.csv$")) |
           # OR matching .xlsx files
           (str_detect(new, "\\.xlsx$") & str_detect(old, "\\.xlsx$"))) %>%
  select(files_std, new, old)


compare_files <- function(file_new, file_old) {
  if(basename(file_new) %>% str_detect("\\.csv$")) {
    data_new <- fread(file_new)
  } 
  if(basename(file_new) %>% str_detect("\\.xlsx$")) {
    data_new <- readxl::read_xlsx(file_new, skip=1)
  }
  if(basename(file_old) %>% str_detect("\\.csv$")) {
    data_old <- fread(file_old)
  }
  if(basename(file_old) %>% str_detect("\\.xlsx$")) {
    data_old <- readxl::read_xlsx(file_old, skip=1)
  }

  
  list(
    file_name_new = basename(file_new),
    file_name_old = basename(file_old),
    new_rows = nrow(data_new),
    old_rows = nrow(data_old),
    row_diff = nrow(data_new) - nrow(data_old),
    new_cols = colnames(data_new),
    old_cols = colnames(data_old),
    # single col value will create a vector column - >1 cols will create list-col
    added_cols = if (length(setdiff(colnames(data_new),
                                    colnames(data_old))) > 0) {
      setdiff(colnames(data_new), colnames(data_old)) 
    } else { NA_character_},
    removed_cols = if (length(setdiff(colnames(data_old),
                                      colnames(data_new))) > 0) {
      setdiff(colnames(data_old), colnames(data_new)) 
    } else {NA_character_},
                   
    data_type_changes = tibble(
      column = union(colnames(data_new), colnames(data_old)),
      # if col is in data_old get its class - else it is NA_character_
      old_type = map_chr(union(colnames(data_new), colnames(data_old)), 
                         ~ ifelse(.x %in% colnames(data_old), 
                                  class(data_old[[.x]])[1], 
                                  NA_character_)),
      # if col is in data_new get its class - else it is NA_character_
      new_type = map_chr(union(colnames(data_new), 
                               colnames(data_old)), 
                         ~ ifelse(.x %in% colnames(data_new), 
                                  class(data_new[[.x]])[1], 
                                  NA_character_))
    ) %>% 
      mutate(type_changed = old_type != new_type)
  )
}

data_level_log <- comparison_pairs %>%
  mutate(comparison = map2(new, old, ~ compare_files(.x, .y)))

# Extract data-level comparison details
data_change_log <- data_level_log %>%
  unnest_wider(comparison) %>%
  select(file_name_new, file_name_old, new_rows, old_rows, row_diff, 
         added_cols, removed_cols, data_type_changes)


```

# Summary

The purpose of this report is to examine and document changes 
between raw input data versions ingested into the ARTIS (Aquatic Resource Trade 
in Species) Model. These raw data will then be processed and standardized to the 
model inputs (data build) fed directly into the ARTIS model. 

### ARTIS model versions

- ARTIS ``r artis_old`` was the previous model run
- ARTIS ``r artis_new`` is upcoming model run

# FAO Global Production

### Raw input data versions

- `r raw_data_type` ``r data_v_old`` is the previous data version
- `r raw_data_type` ``r data_v_new`` is the data version to be ingested in the upcoming ARTIS ``r artis_new`` model run

### File changes

```{r logic-files-missing}
#| results: 'asis'
#| include: TRUE

if (any(file_level_log$exists_in_old == FALSE)) {
  missing_files <- file_level_log |>
    filter(!exists_in_old) |>
    pull(new) |>
    basename() |>
    paste(collapse = ", ")
  
  cat(paste0(
    "The new ", data_v_new, " data version contains files that are not present in the old ",
    data_v_old, " version: ", missing_files
  ))
}
```

```{r logic-files-added}
#| results: 'asis'
#| include: TRUE

if (any(file_level_log$exists_in_new == FALSE)) {
  cat(paste0("The old ", data_v_old, " data version contains files that are not present in the new ", data_v_new, " version: ", basename(file_level_log[[which(file_level_log$exists_in_new == FALSE), "old"]])
                 )
          )
}
```

### Column changes

Only matching .csv and .xlsx files were compared.

```{r}
#| label: tab:added_removed_columns
#| include: TRUE
#| results: 'asis'

# Filter the data_change_log to only include files with added or removed columns
added_cols_log <- data_change_log %>% 
  filter(!is.na(added_cols) | !is.na(removed_cols)) %>% 
  select(file_name_new, added_cols, removed_cols)

if (nrow(added_cols_log) > 0 ) {
  # Create a kableExtra table with automatic scaling
  added_cols_log %>%
    kbl(
      col.names = c("File Name", "Added Columns", "Removed Columns"),
      caption = "Files where columns were added/removed in the new version.",
      align = c('l', 'c', 'c')
    ) %>%
    kable_styling(
      full_width = TRUE, 
      latex_options = c("scale_down", "hold_position")
    )
}

```

### Sciname

```{r sciname-diff-raw}
#| include: TRUE
#| results: 'asis'

# get old production species - uncleaned
raw_prod_old_fn <- glue("GlobalProduction_{data_v_old}.zip")
raw_prod_old_sp <- artis::rebuild_fao_2023_dat(
  datadir = datadir_old, 
  filename = raw_prod_old_fn) %>%
  select(
    sciname = species_scientific_name, 
    common_name = species_name_en) %>%
  mutate(
    sciname = tolower(sciname),
    common_name = tolower(common_name)) %>%
  distinct()

# get new production species - uncleaned
raw_prod_new_fn <- glue("GlobalProduction_{data_v_new}.zip")
raw_prod_new_sp <- artis::rebuild_fao_2023_dat(
  datadir = datadir_new, 
  filename = raw_prod_new_fn) %>%
  select(
    sciname = species_scientific_name, 
    common_name = species_name_en) %>%
  mutate(
    sciname = tolower(sciname),
    common_name = tolower(common_name)) %>%
  distinct()

raw_prod_old_sp_only <- anti_join(raw_prod_old_sp, raw_prod_new_sp, by = "sciname") %>%
  arrange(sciname) %>% 
  mutate(version = data_v_old)

raw_prod_new_sp_only <- anti_join(raw_prod_new_sp, raw_prod_old_sp, by = "sciname") %>%
  arrange(sciname)  %>% 
  mutate(version = data_v_new)

raw_prod_new_sp_only %>%
  kable(
    caption = glue("Taxa unique to {data_v_new} FAO production data"),
    booktabs = TRUE,
    col.names = c("Scientific Name", "Common Name", "version")
  ) %>%
  kable_styling(font_size = 8)   # set size in points

raw_prod_old_sp_only %>%
  kable(
    caption = glue("Taxa unique to {data_v_old} FAO production data"),
    booktabs = TRUE,
    col.names = c("Scientific Name", "Common Name", "version")
  ) %>%
  kable_styling(font_size = 8)   # set size in points

```



# BACI Bilateral Trade

```{r}
path_baci_old <- file.path(datadir_old, "baci_raw")
path_baci_new <- file.path(datadir_new, "baci_raw")
```

### Countries

```{r}
# Old BACI assessment ----------------------------------

# Get HS version from raw baci data directories
hs_versions_old <- list.files(
  path = path_baci_old,
  pattern = "HS\\d{2}",
  recursive = FALSE
) |>
  str_subset("\\.zip$", negate = TRUE) |>
  str_extract("HS\\d{2}")

countries_old <- list.files(
  path = path_baci_old,
  pattern = "country_codes",
  recursive = TRUE
) 
# Check assumption that all country_code_<baci_version..csv files are identical
countries_old_data <- file.path(path_baci_old, countries_old) |>
  map(fread)

all_identical <- countries_old_data |>
  map_lgl(\(x) identical(as.data.frame(x), as.data.frame(countries_old_data[[1]]))) |>
  all()

all_identical

# log in the validation log
validation_log[[glue("BACI {baci_v_old} country code files identical across HS versions")]] <- list(
  passed = all_identical,
  notes  = "Check for differing file versions in path_baci_old. Files should all share the same V-version suffix."
)

# New Baci assessment ------------------------------
# Get HS version from raw baci data directories
hs_versions_new <- list.files(
  path = path_baci_new,
  pattern = "HS\\d{2}",
  recursive = FALSE
) |>
  str_subset("\\.zip$", negate = TRUE) |>
  str_extract("HS\\d{2}")

countries_new <- list.files(
  path = path_baci_new,
  pattern = "country_codes",
  recursive = TRUE
) 
# Check assumption that all country_code_<baci_version..csv files are identical
countries_new_data <- file.path(path_baci_new, countries_new) |>
  map(fread)

all_identical <- countries_new_data |>
  map_lgl(\(x) identical(as.data.frame(x), as.data.frame(countries_new_data[[1]]))) |>
  all()

all_identical

# log in the validation log
validation_log[[glue("BACI {baci_v_new} country code files identical across HS versions")]] <- list(
  passed = all_identical,
  notes  = "Check for differing file versions in path_baci_new. Files should all share the same V-version suffix."
)
```

### Products
```{r}
#| include: TRUE

# Get all product_codes files for old and new BACI
products_old <- list.files(
  path = path_baci_old,
  pattern = "product_codes",
  recursive = TRUE
)

products_new <- list.files(
  path = path_baci_new,
  pattern = "product_codes",
  recursive = TRUE
)

# Build tidy tibbles with HS version extracted, then join to pair by HS version
products_old_df <- tibble(rel_path = products_old) |>
  mutate(
    hs_version  = str_extract(rel_path, "HS\\d{2}"),
    full_path   = file.path(path_baci_old, rel_path),
    baci_v      = baci_v_old
  )

products_new_df <- tibble(rel_path = products_new) |>
  mutate(
    hs_version  = str_extract(rel_path, "HS\\d{2}"),
    full_path   = file.path(path_baci_new, rel_path),
    baci_v      = baci_v_new
  )

# Pair old and new by HS version
products_paired <- inner_join(
  products_old_df, 
  products_new_df,
  by = join_by(hs_version),
  suffix = c("_old", "_new")
)

# Compare each HS version pair
compare_product_codes <- function(path_old, path_new, hs_ver) {
  old_dat <- fread(path_old)
  new_dat <- fread(path_new)

  codes_removed <- anti_join(
    as.data.frame(old_dat), as.data.frame(new_dat), by = "code"
  )
  codes_added <- anti_join(
    as.data.frame(new_dat), as.data.frame(old_dat), by = "code"
  )

  tibble(
    hs_version      = hs_ver,
    n_codes_old     = nrow(old_dat),
    n_codes_new     = nrow(new_dat),
    n_codes_removed = nrow(codes_removed),
    n_codes_added   = nrow(codes_added),
    codes_removed   = list(codes_removed),
    codes_added     = list(codes_added)
  )
}

product_code_comparison <- products_paired |>
  pmap(
    \(hs_version, full_path_old, full_path_new, ...) 
      compare_product_codes(full_path_old, full_path_new, hs_version)
  ) |>
  list_rbind()

# Log validation checks per HS version
walk(product_code_comparison$hs_version, \(hs_ver) {
  row <- filter(product_code_comparison, hs_version == hs_ver)
  passed <- row$n_codes_removed == 0 & row$n_codes_added == 0
  validation_log[[glue("BACI {hs_ver} product codes unchanged {baci_v_old} -> {baci_v_new}")]] <<- list(
    passed = passed,
    notes  = glue("Check product_codes_{hs_ver}_*.csv files in path_baci_old and path_baci_new. {row$n_codes_added} codes added, {row$n_codes_removed} codes removed.")
  )
})

# Display product code comparison summary table
product_code_comparison |>
  select(hs_version, n_codes_old, n_codes_new, n_codes_removed, n_codes_added) |>
  gt() |>
  tab_header(title = "BACI Product Code Changes by HS Version",
             subtitle = glue("{baci_v_old} vs {baci_v_new}")) |>
  cols_label(
    hs_version      = "HS Version",
    n_codes_old     = glue("# Codes ({baci_v_old})"),
    n_codes_new     = glue("# Codes ({baci_v_new})"),
    n_codes_removed = "# Removed",
    n_codes_added   = "# Added"
  ) |>
  tab_style(
    style = cell_fill(color = "#f4cccc"),
    locations = cells_body(rows = n_codes_removed > 0 | n_codes_added > 0)
  )
```

# Validation Summary

```{r validation-summary}
#| include: TRUE

validation_log |>
  tibble::enframe(name = "check", value = "result") |>
  mutate(
    passed = map_lgl(result, "passed"),
    notes  = map_chr(result, "notes"),
    status = if_else(passed, "✅ PASS", "❌ FAIL"),
    notes  = if_else(passed, "—", notes)
  ) |>
  select(check, status, notes) |>
  gt() |>
  tab_header(title = "Raw Data Validation Summary") |>
  cols_label(
    check  = "Check",
    status = "Status",
    notes  = "Next Steps if Failing"
  ) |>
  tab_style(
    style = cell_fill(color = "#f4cccc"),
    locations = cells_body(rows = status == "❌ FAIL")
  ) |>
  tab_style(
    style = cell_fill(color = "#d9ead3"),
    locations = cells_body(rows = status == "✅ PASS")
  ) |>
  cols_width(
    check ~ px(250),
    notes ~ px(280)
  ) |>
  tab_options(table.font.size = 10)
```