---
title: "ARTIS Raw Data Structure Assessment"
subtitle: "Phase 1: Schema and File Structure Changes"
author: "Althea Marks"
format:
  pdf:
    papersize: letter
execute:
  echo: false
  include: false
  message: false
  warning: false
---

```{r setup}
#### Libraries #####
library(data.table)
library(dplyr)
library(readr)
library(stringr)
library(purrr)
library(tidyr)
library(knitr)
library(kableExtra)
library(readxl)
library(glue)

#### Directory setup #####
# Newest data
datadir_new <- file.path("~", "Documents", "UW-SAFS", "ARTIS", "data", "model_inputs_raw_3.0")
# Previous data ingest
datadir_old <- file.path("~", "Documents", "UW-SAFS", "ARTIS", "data", "model_inputs_raw_2.0")

# Version identifiers for report
artis_old <- "v2.0"
artis_new <- "v3.0"
data_v_old <- "2025.1.0"
data_v_new <- "2026.1.0"
```

```{r data-source-inventory}
# Define data source patterns and their assessment functions
data_sources <- list(
  fao_production = list(
    pattern = regex("(global[\\s_-]*production)", ignore_case = TRUE),
    key_files = c("Global_production_quantity.csv", "Global_production_quantity.xlsx"),
    description = "FAO Global Production"
  ),
  baci = list(
    pattern = regex("(baci)", ignore_case = TRUE),
    key_files = c(),  # Will be detected automatically
    description = "BACI Trade Data"
  ),
  hs_codes = list(
    pattern = "ALL_HS_Codes.csv",
    key_files = c("ALL_HS_Codes.csv"),
    description = "HS Code Classifications"
  )
)

# Function to scan and categorize files by data source
assess_data_source <- function(datadir_new, datadir_old, source_config) {
  # Get subdirectories
  subdirs_new <- list.dirs(datadir_new, recursive = TRUE, full.names = TRUE)
  subdirs_old <- list.dirs(datadir_old, recursive = TRUE, full.names = TRUE)
  
  if (is.character(source_config$pattern)) {
    # Direct file matching (like ALL_HS_Codes.csv)
    files_new <- list.files(datadir_new, pattern = source_config$pattern, 
                           recursive = TRUE, full.names = TRUE)
    files_old <- list.files(datadir_old, pattern = source_config$pattern, 
                           recursive = TRUE, full.names = TRUE)
  } else {
    # Directory pattern matching
    matching_dirs_new <- subdirs_new[str_detect(subdirs_new, source_config$pattern)]
    matching_dirs_old <- subdirs_old[str_detect(subdirs_old, source_config$pattern)]
    
    files_new <- if(length(matching_dirs_new) > 0) {
      list.files(matching_dirs_new, recursive = TRUE, full.names = TRUE)
    } else { character(0) }
    
    files_old <- if(length(matching_dirs_old) > 0) {
      list.files(matching_dirs_old, recursive = TRUE, full.names = TRUE)
    } else { character(0) }
  }
  
  list(
    source = source_config$description,
    files_new = files_new,
    files_old = files_old,
    dirs_new = if(exists("matching_dirs_new")) matching_dirs_new else character(0),
    dirs_old = if(exists("matching_dirs_old")) matching_dirs_old else character(0)
  )
}

# Assess all data sources
source_assessments <- map(data_sources, ~assess_data_source(datadir_new, datadir_old, .x))
names(source_assessments) <- names(data_sources)
```

```{r file-structure-comparison}
# Universal file comparison function - ROBUST FIX
create_file_comparison <- function(files_new, files_old, source_name) {
  if(length(files_new) == 0 && length(files_old) == 0) {
    return(tibble(
      source = source_name,
      files_std = character(0),
      new = character(0),
      old = character(0),
      exists_in_old = logical(0),
      exists_in_new = logical(0),
      size_old_mb = numeric(0),
      size_new_mb = numeric(0),
      size_change_mb = numeric(0)
    ))
  }
  
  all_files <- tibble(
    source = source_name,
    full_path = c(files_new, files_old),
    files = c(basename(files_new), basename(files_old)),
    directory = c(rep("new", length(files_new)), rep("old", length(files_old)))
  ) %>%
    mutate(
      files_std = files %>% 
        str_remove("^filtered_") %>%
        str_remove("_V\\d{6,7}[a-z]*") %>%
        str_replace_all("\\.[a-zA-Z0-9]+$", "") %>%
        str_replace_all("[^a-zA-Z0-9]", "_") %>%
        str_to_lower()
    )
  
  file_level_log <- all_files %>%
    pivot_wider(
      id_cols = c(source, files_std),
      names_from = directory, 
      values_from = full_path
    ) %>%
    mutate(
      exists_in_old = !is.na(old),
      exists_in_new = !is.na(new)
    ) %>%
    rowwise() %>%
    mutate(
      size_old_mb = if(exists_in_old && !is.na(old) && file.exists(old)) {
        file.info(old)$size / 1e6
      } else {
        NA_real_
      },
      size_new_mb = if(exists_in_new && !is.na(new) && file.exists(new)) {
        file.info(new)$size / 1e6
      } else {
        NA_real_
      }
    ) %>%
    ungroup() %>%
    mutate(
      size_change_mb = size_new_mb - size_old_mb
    )
  
  return(file_level_log)
}
```

```{r schema-comparison}
# Enhanced schema comparison function
compare_schemas <- function(file_new, file_old, max_rows_sample = 1000) {
  read_file_smart <- function(filepath, max_rows = max_rows_sample) {
    ext <- tools::file_ext(filepath)
    if(ext == "csv") {
      # Sample first N rows for large files to check schema
      if(file.info(filepath)$size > 50*1024*1024) { # If > 50MB
        fread(filepath, nrows = max_rows)
      } else {
        fread(filepath)
      }
    } else if(ext %in% c("xlsx", "xls")) {
      read_excel(filepath, n_max = max_rows)
    } else {
      return(NULL)
    }
  }
  
  tryCatch({
    data_new <- read_file_smart(file_new)
    data_old <- read_file_smart(file_old)
    
    if(is.null(data_new) || is.null(data_old)) {
      return(NULL)
    }
    
    list(
      file_name_new = basename(file_new),
      file_name_old = basename(file_old),
      new_rows = nrow(data_new),
      old_rows = nrow(data_old),
      row_diff = nrow(data_new) - nrow(data_old),
      row_diff_pct = round(((nrow(data_new) - nrow(data_old)) / nrow(data_old)) * 100, 2),
      new_cols = ncol(data_new),
      old_cols = ncol(data_old),
      col_diff = ncol(data_new) - ncol(data_old),
      new_col_names = list(colnames(data_new)),
      old_col_names = list(colnames(data_old)),
      added_cols = list(setdiff(colnames(data_new), colnames(data_old))),
      removed_cols = list(setdiff(colnames(data_old), colnames(data_new))),
      data_types_new = list(sapply(data_new, function(x) class(x)[1])),
      data_types_old = list(sapply(data_old, function(x) class(x)[1]))
    )
  }, error = function(e) {
    list(
      file_name_new = basename(file_new),
      file_name_old = basename(file_old),
      error_msg = conditionMessage(e)
    )
  })
}

# Apply schema comparison to matching files
comparable_files <- all_file_comparisons %>%
  filter(exists_in_old & exists_in_new) %>%
  filter(!is.na(new), !is.na(old))

schema_comparisons <- comparable_files %>%
  filter(nrow(.) > 0) %>%
  rowwise() %>%
  mutate(
    schema_comparison = list(compare_schemas(new, old))
  ) %>%
  ungroup()

if(nrow(schema_comparisons) > 0) {
  schema_details <- schema_comparisons %>%
    select(source, files_std, schema_comparison) %>%
    unnest_wider(schema_comparison)
} else {
  schema_details <- tibble()
}
```

# Executive Summary

This Phase 1 assessment evaluates **structural changes** in raw data sources that could break the ARTIS cleaning pipeline. It focuses on detecting schema changes, file additions/removals, and data type modifications **before** applying any data cleaning logic.

**Purpose**: Identify breaking changes that require updates to `01-clean-input-data.R` before proceeding with content-level analysis.

**Data Version Comparison**:
- Previous: ARTIS `r artis_old` (data version `r data_v_old`)
- Current: ARTIS `r artis_new` (data version `r data_v_new`)

# Data Source Inventory

```{r source-summary}
#| include: true
#| results: asis

source_summary <- map_dfr(source_assessments, ~tibble(
  source = .x$source,
  files_new_count = length(.x$files_new),
  files_old_count = length(.x$files_old),
  dirs_new_count = length(.x$dirs_new),
  dirs_old_count = length(.x$dirs_old)
))

source_summary %>%
  kbl(
    col.names = c("Data Source", "Files (New)", "Files (Old)", "Dirs (New)", "Dirs (Old)"),
    caption = "Raw data source inventory",
    booktabs = TRUE
  ) %>%
  kable_styling(
    full_width = TRUE,
    latex_options = c("hold_position", "scale_down")
  )
```

# File-Level Changes

## New and Missing Files

```{r file-changes-summary}
#| include: true
#| results: asis

file_changes <- all_file_comparisons %>%
  group_by(source) %>%
  summarise(
    total_files = n(),
    files_added = sum(!exists_in_old, na.rm = TRUE),
    files_removed = sum(!exists_in_new, na.rm = TRUE),
    files_unchanged = sum(exists_in_old & exists_in_new, na.rm = TRUE),
    .groups = "drop"
  )

file_changes %>%
  kbl(
    col.names = c("Data Source", "Total Files", "Added", "Removed", "Unchanged"),
    caption = "File-level changes by data source",
    booktabs = TRUE
  ) %>%
  kable_styling(
    full_width = TRUE,
    latex_options = c("hold_position", "scale_down")
  )
```

## File Size Changes

```{r file-size-changes}
#| include: true
#| results: asis

significant_size_changes <- all_file_comparisons %>%
  filter(exists_in_old & exists_in_new) %>%
  filter(!is.na(size_change_mb)) %>%
  mutate(
    size_change_pct = round((size_change_mb / size_old_mb) * 100, 1)
  ) %>%
  filter(abs(size_change_pct) > 10) %>% # Only show > 10% changes
  arrange(desc(abs(size_change_pct)))

if(nrow(significant_size_changes) > 0) {
  significant_size_changes %>%
    select(source, files_std, size_old_mb, size_new_mb, size_change_mb, size_change_pct) %>%
    kbl(
      col.names = c("Source", "File", "Old (MB)", "New (MB)", "Change (MB)", "Change (%)"),
      caption = "Significant file size changes (>10%)",
      booktabs = TRUE,
      digits = 2
    ) %>%
    kable_styling(
      full_width = TRUE,
      latex_options = c("hold_position", "scale_down")
    )
}
```

# Schema-Level Changes

## Row Count Changes

```{r row-changes}
#| include: true
#| results: asis

if(nrow(schema_details) > 0 && "row_diff_pct" %in% names(schema_details)) {
  row_changes <- schema_details %>%
    filter(!is.na(row_diff_pct)) %>%
    filter(abs(row_diff_pct) > 5) %>% # Only show > 5% changes
    select(source, file_name_new, old_rows, new_rows, row_diff, row_diff_pct) %>%
    arrange(desc(abs(row_diff_pct)))
  
  if(nrow(row_changes) > 0) {
    row_changes %>%
      kbl(
        col.names = c("Source", "File", "Old Rows", "New Rows", "Difference", "Change (%)"),
        caption = "Significant row count changes (>5%)",
        booktabs = TRUE
      ) %>%
      kable_styling(
        full_width = TRUE,
        latex_options = c("hold_position", "scale_down")
      )
  }
}
```

## Column Schema Changes

```{r column-changes}
#| include: true
#| results: asis

if(nrow(schema_details) > 0) {
  column_changes <- schema_details %>%
    filter(!is.na(col_diff) & col_diff != 0) %>%
    select(source, file_name_new, old_cols, new_cols, col_diff)
  
  if(nrow(column_changes) > 0) {
    column_changes %>%
      kbl(
        col.names = c("Source", "File", "Old Columns", "New Columns", "Difference"),
        caption = "Files with column count changes",
        booktabs = TRUE
      ) %>%
      kable_styling(
        full_width = TRUE,
        latex_options = c("hold_position", "scale_down")
      )
  }