---
title: "07-post-processing-validation"
author: "Jessica Gephart"
date: "2025-06-26"
output: html_document
---

# Overview

The Aquatic Resource Trade in Species (ARTIS) model generates two primary outputs: the taxa disaggregated trade record file and the taxa disaggregated consumption file. The trade file represents a disaggregation of reported trade records, which reflect individual trade transactions such that a product flowing from country A to country B and then to country C would be represented as two trade records/observations (A exporting to B and B exporting to C). Meanwhile, the apparent consumption file represents a linear flow of products wherein the the ARTIS trade file is used with aligned production data to calculate apparent consumption based on a "disappearance" model of products. Aquatic resources imported for consumption are disaggregated into taxa, production methods, and exporter/source country. 

In addition to the trade and consumption files, we output several additional files that users may find useful:  

* prod.csv - this file represents the production data used within the ARTIS model, ensuring that the principle of mass-balance is maintained. All taxa names and country boundaries have been standardized to align with ARTIS. This can join ARTIS trade or consumption by year, source_country_iso3c and sciname. 
* baci.csv - this file represents the trade data in terms of hs product codes used within the ARTIS model. This file is primarily useful within this validation script to confirm that all trade in ARTIS adds back up to the input trade values. This can join ARTIS trade by hs_version, year, importer, exporter, hs6 after ARTIS trade has been summarized by those columns. 
* sciname_metadata.csv - this file includes taxonomic information, common names, and standard taxa groups that are often useful for exploring and presenting the data. This can join ARTIS trade or consumption by sciname or sciname_max_resolved.
* code_max_resolved_taxa.csv - this file can be joined to ARTIS to add more specific taxa names when the taxonomic resolution of a code is finer than the information from the production data (e.g., actinopterygii within a code where all species belong to thunnus). This can join ARTIS trade by hs_version, hs6, and sciname. Note that sciname_max_resolved will not add back up to the production data when grouped by source country. 
* products.csv - this file includes the hs code descriptions and it can be joined to ARTIS trade data by hs_version and hs6 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, 
                      fig.width = 6, fig.height = 4)
```

```{r load packages}
library(arrow)
library(data.table)
library(tidyverse)
library(exploreARTIS)
```

```{r load data}
# Set database folder location
db_folder <- "/Users/gephart/Library/CloudStorage/Dropbox/Research/ARTIS/ARTIS_1.1.0_FAO_2025-06-10/"

# load ARTIS data set
# point to the dataset folder
artis_ds <- arrow::open_dataset(file.path(db_folder, "datasets", "artis_parquet_dataset"))

# custom timeseries filter - result "artis_ts_result" is NOT in-memory
# this is a "Predicate Pushdown" or "On-disk filtering" method. 
# Only able to filter or subset data - no manipulation until data is in-memory

artis_ts_result <- artis_ds %>%
  filter(
    (hs_version == "HS96" & year <= 2003) |
      (hs_version == "HS02" & year >= 2004 & year <= 2009) |
      (hs_version == "HS07" & year >= 2010 & year <= 2012) |
      (hs_version == "HS12" & year >= 2013 & year <= 2020)
  ) # %>% Can add more data filtering

# Bring the results into R memory by adding "collect()" at the end of the pipe
artis_ts_df <- artis_ts_result %>% 
  collect()

# create table to compare with baci
artis_hs_summary <- artis_ds %>%
  group_by(hs_version, year, exporter_iso3c, importer_iso3c, hs6) %>%
  summarise(product_weight_t = sum(product_weight_t, na.rm = TRUE))

artis_hs_summary <- artis_hs_summary %>%
  collect() 

# load consumption data set
# # point to the dataset folder
artis_ds <- arrow::open_dataset(file.path(db_folder, "datasets", "consumption_parquet_dataset"))

artis_ts_result <- artis_ds %>%
  filter(
    # Remove small volumes to allow the file to render efficiently
    #consumption_live_t >= 1,
    (hs_version == "HS96" & year <= 2003) |
      (hs_version == "HS02" & year >= 2004 & year <= 2009) |
      (hs_version == "HS07" & year >= 2010 & year <= 2012) |
      (hs_version == "HS12" & year >= 2013 & year <= 2020)
  ) # %>% Can add more data filtering

# Bring the results into R memory by adding "collect()" at the end of the pipe
consumption_ts_df <- artis_ts_result %>%
  collect()

# Load attribute tables
prod <- fread(file.path(db_folder, "attribute-tables", "prod.csv"))
baci <- fread(file.path(db_folder, "attribute-tables", "baci.csv")) 
sciname <- fread(file.path(db_folder, "attribute-tables", "sciname_metadata.csv"))
code_max_resolved_taxa <- fread(file.path(db_folder, "attribute-tables", "code_max_resolved_taxa.csv"))

```

```{r generate parameters}
max_year <- max(artis_ts_df$year)
```

# Results
## Global trade summary stats

```{r global stats}
total_1996 <- artis_ts_df %>%
  filter(year == 1996) %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_max_year <- artis_ts_df %>%
  filter(year == max_year) %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_capture_1996 <- artis_ts_df %>%
  filter(year == 1996, method == "capture") %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_capture_max_year <- artis_ts_df %>%
  filter(year == max_year, method == "capture") %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_aquaculture_1996 <- artis_ts_df %>%
  filter(year == 1996, method == "aquaculture") %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_aquaculture_max_year <- artis_ts_df %>%
  filter(year == max_year, method == "aquaculture") %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_dom_1996 <- artis_ts_df %>%
  filter(year == 1996, dom_source == "domestic") %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_dom_max_year <- artis_ts_df %>%
  filter(year == max_year, dom_source == "domestic") %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_error_1996 <- artis_ts_df %>%
  filter(year == 1996, dom_source == "error") %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)

total_error_max_year <- artis_ts_df %>%
  filter(year == max_year, dom_source == "error") %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  pull(live_weight_t)
```


* Global trade increased from `r round(total_1996/1000000)` mil t (live weight) in 1996 to `r round(total_max_year/1000000)` mil t in `r paste(max_year)`, a `r round(100*(total_max_year-total_1996)/total_1996)`% increase. 
* Capture production represented `r round(100*total_capture_max_year/total_max_year)`% in `r paste(max_year)`, compared to `r round(100*total_capture_1996/total_1996)`% in 1996. Conversely, the share of exports from aquaculture increased from `r round(100*total_aquaculture_1996/total_1996)`% in 1996 to `r round(100*total_aquaculture_max_year/total_max_year)`% in `r paste(max_year)`. 
* In `r paste(max_year)`, the share of exports from domestic production was `r round(100*total_dom_max_year/total_max_year)`%, compared to `r round(100*total_dom_1996/total_1996)`% in 1996.
* Globally, `r round(100*total_error_max_year/total_max_year)`% of exports could not be explained by production and imports in `r paste(max_year)`.

## Method and habitat sourcing
```{r method-habitat figs}
artis_ts_df %>%
  mutate(habitat_method = paste(habitat, method)) %>%
  plot_ts(artis_var = "habitat_method", plot.type = "stacked", 
          prop_flow_cutoff = 0,
          plot.title = "Global trade by production source")
```

## Production sourcing of exports
```{r export source figs}
artis_ts_df %>%
  plot_ts(artis_var = "dom_source", plot.type = "stacked", 
          prop_flow_cutoff = 0, 
          plot.title = "Global trade by export source")
```

## Global trade by species
```{r top species figs}
artis_ts_df %>%
  plot_ts(artis_var = "sciname", prop_flow_cutoff = 0.02, plot.type = "stacked",
          plot.title = "Global trade by taxa")

artis_ts_df %>%
  filter(year == max_year) %>%
  plot_bar(bar_group = "sciname", 
           plot.title = paste("Top 15 traded species,", max_year), 
           top_n = 15)
```

## Top exporters 
```{r top exporters figs}
artis_ts_df %>%
  plot_ts(artis_var = "exporter_iso3c", plot.type = "stacked", 
          prop_flow_cutoff = 0.03, 
          plot.title = "Global trade by exporter")

artis_ts_df %>%
  filter(year == max_year) %>%
  plot_bar(bar_group = "exporter_iso3c", 
           plot.title = paste("Top 15 exporters,", max_year), 
           top_n = 15)
```

## Top importers
```{r top importers figs}
artis_ts_df %>%
  plot_ts(artis_var = "importer_iso3c", plot.type = "stacked", 
          prop_flow_cutoff = 0.03,
          plot.title = "Global trade by importer")

artis_ts_df %>%
  filter(year == max_year) %>%
  plot_bar(bar_group = "importer_iso3c", 
           plot.title = paste("Top 15 importers,", max_year), 
           top_n = 15)
```

## Top consumers
```{r top consumers figs}
consumption_ts_df %>% 
  plot_ts(artis_var = "consumer_iso3c", value = "consumption_live_t", 
          plot.type = "stacked", plot.title = "Global consumption", 
          prop_flow_cutoff = 0.03)

consumption_ts_df %>%
  filter(year == max_year, end_use == "direct human consumption") %>%
  plot_bar(bar_group = "consumer_iso3c", top_n = 15, fill_type = "method",
           value = "consumption_live_t", 
           plot.title = paste("Top 15 consumers, t live weight,", max_year))

consumption_ts_df %>%
  filter(year == max_year) %>%
  plot_bar(bar_group = "consumer_iso3c", top_n = 30, fill_type = "method",
           value = "consumption_percap_live_kg_capped", 
           plot.title = paste("Top 30 consumers, per capita,", max_year))
```

## Consumption sourcing
```{r consumption sourcing figs}
consumption_ts_df %>%
  plot_ts(artis_var = "consumption_source", value = "consumption_live_t", 
          plot.type = "stacked", 
          plot.title = "Global consumption sourcing")
```


# QA testing

## Compare ARTIS trade with BACI
When grouped by year, exporter, importer and HS code, ARTIS trade data should generally sum to BACI in terms of product weight. Volumes are expected to be slightly lower in ARTIS trade data than BACI due to small values dropped within the ARTIS model. 

```{r compare with baci}
# Filter BACI down to the hs codes appearing in ARTIS for a given HS version
baci <- artis_hs_summary %>%
  ungroup() %>%
  select(hs_version, hs6) %>%
  distinct() %>%
  mutate(hs6 = as.numeric(hs6)) %>%
  left_join(baci, by = c("hs_version", "hs6"))

# Join BACI and ARTIS trade to calculate product volume total differences
# by hs version, year, exporter, importer and hs6 code
baci_check <- artis_hs_summary %>%
  rename("artis_product_weight_t" = "product_weight_t") %>%
  mutate(hs6 = as.numeric(hs6)) %>%
  full_join(baci %>%
              rename("baci_product_weight_t" = "product_weight_t"), 
            by = c("hs_version", "year", 
                         "exporter_iso3c", "importer_iso3c", "hs6")) %>% 
  mutate(diff = baci_product_weight_t - artis_product_weight_t) %>% 
  arrange(desc(diff))

baci_check_na <- baci_check %>%
  filter(is.na(diff))
```

We observe that the maximum difference between BACI and ARTIS is `r max(baci_check$diff, na.rm=TRUE)` and `r baci_check %>% filter(diff < -0.0001) %>% nrow()` observations where ARTIS product totals are greater than BACI (no product mass should be gained). The join produces `r nrow(baci_check_na)` NAs (`r 100*nrow(baci_check_na)/nrow(baci)`% of rows in BACI), which represent flows without a match in ARTIS, and have a maximum volume of `r max(baci_check_na$baci_product_weight_t)`t (`r 100*(baci_check_na %>% filter(baci_product_weight_t <= 1) %>% nrow())/nrow(baci_check_na)` are 1t or less). Note that small volumes are dropped throughout the ARTIS model to improve computational efficiency.  

```{r}
# Calculate totals for all hs version-year combinations
baci_ts_plot <- baci_check %>%
  group_by(hs_version, year) %>%
  summarise(artis_product_weight_t = sum(artis_product_weight_t, na.rm = TRUE),
            baci_product_weight_t = sum(baci_product_weight_t, na.rm = TRUE)) 

# Calculate totals for custom artis time series
custom_artis_ts_plot <- artis_ts_df %>%
  group_by(hs_version, year) %>%
  summarise(product_weight_t = sum(product_weight_t))

ggplot() +
  geom_line(data = baci_ts_plot, aes(x = year, y = baci_product_weight_t, 
                                     color = hs_version), 
            linewidth = 1) +
  geom_line(data = baci_ts_plot, aes(x = year, y = artis_product_weight_t, 
                                     color = hs_version),
            linetype = "dashed", linewidth = 1) + 
  geom_line(data = custom_artis_ts_plot, aes(x = year, y = product_weight_t),
            color = "black", linewidth = 0.5) +
  labs(x = "", y = "Global trade (product weight, t)", color = "HS version") +
  theme_bw()
  
```

## Compare ARTIS trade and consumption

```{r}
foreign_consumption_totals <- consumption_ts_df %>%
  filter(consumption_source != "domestic") %>%
  group_by(year, source_country_iso3c, sciname) %>%
  summarise(consumption_live_t = sum(consumption_live_t)) %>%
  ungroup()

domestic_export_totals <- artis_ts_df %>%
  filter(dom_source == "domestic") %>%
  group_by(year, source_country_iso3c, sciname) %>%
  summarise(live_weight_t = sum(live_weight_t)) %>%
  ungroup()

trade_consumption_check <- foreign_consumption_totals %>%
  full_join(domestic_export_totals, 
            by = c("year", "source_country_iso3c", "sciname")) %>%
  mutate(diff = consumption_live_t-live_weight_t)
```

Foreign consumption totals should sum to domestic exports in ARTIS trade. There are `r trade_consumption_check %>% filter(diff < -0.0001) %>% nrow()` cases where foreign consumption is less than the total domestic exports of a species from a given country and `r trade_consumption_check %>% filter(diff > 10) %>% nrow()` cases (out of `r nrow(trade_consumption_check)`, or `r round(100*(trade_consumption_check %>% filter(diff > 10) %>% nrow())/nrow(trade_consumption_check), 2)`% of cases) where estimated consumption is at least 10t greater than domestic exports of a species from a given country. 

## Compare ARTIS with production

```{r compare with production}
# Confirm that domestic exports of a given species do not exceed production
compare_trade_prod <- artis_ts_df %>%
  filter(dom_source == "domestic") %>%
  group_by(year, source_country_iso3c, sciname, habitat, method) %>%
  summarise(trade_live_weight_t = sum(live_weight_t)) %>%
  left_join(prod %>%
              group_by(year, iso3c, sciname, habitat, method) %>%
              summarise(live_weight_t = sum(live_weight_t)), 
            by = c("year", "source_country_iso3c" = "iso3c", 
                   "sciname", "habitat", "method")) %>%
  mutate(diff = live_weight_t - trade_live_weight_t) %>%
  filter(diff < -0.0001)

# Confirm all production is consumed
compare_consumption_prod <- consumption_ts_df %>% 
  group_by(year, source_country_iso3c, sciname, habitat, method) %>% 
  summarise(consumption_live_t = sum(consumption_live_t)) 

compare_consumption_prod_nrow <- nrow(compare_consumption_prod)

compare_consumption_prod <- compare_consumption_prod %>%
  full_join(prod %>%
              group_by(year, iso3c, sciname, habitat, method) %>%
              summarise(live_weight_t = sum(live_weight_t)),
            by = c("year", "source_country_iso3c" = "iso3c", "sciname", 
                   "habitat", "method")) %>%
  mutate(diff = abs(consumption_live_t - live_weight_t), 
         percent_diff = 100*(consumption_live_t - live_weight_t)/live_weight_t)

large_diff <- compare_consumption_prod %>% 
  filter(diff > 10)

```

Domestic exports in ARTIS should not exceed production in the trade.  There are `r nrow(compare_trade_prod)` instances where domestic exports of a species by a country exceed production in that country for that year. 

Additionally, when grouped by source country, consumption should add back up to production (i.e., all production should be consumed by someone). There are `r compare_consumption_prod %>% filter(diff < -0.0001) %>% nrow()` cases where foreign consumption is less than the corresponding production of a species from a given country and `r compare_consumption_prod %>% filter(diff > 10) %>% nrow()` cases (out of `r nrow(compare_consumption_prod)`, or `r round(100*(compare_consumption_prod %>% filter(diff > 10) %>% nrow())/nrow(compare_consumption_prod), 2)`%) where estimated consumption is at least 10t greater than the corresponding production of a species from a given country. 

## Attribute table checks
```{r attribute checks}
# For each attribute table, check for NA's in file and then join to ARTIS files 
# to confirm there are no NA's generated
```

## Confirm joins with attribute tables
Check for any NA's in joins with attribute tables. 
```{r attribute joins}
# Description of attribute tables and join columns are at the top of this doc
```

